---
title: "FIAwrapper: Complete Package Guide"
subtitle: "Clean FIA Tree Lists in Forestry Language"
author: "Ryan Smith"
date: last-modified
format:
  html:
    toc: true
    toc-depth: 4
    toc-expand: 2
    number-sections: true
    code-fold: false
    code-overflow: wrap
    theme: flatly
    highlight-style: github
    anchor-sections: true
    smooth-scroll: true
    link-external-newwindow: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  eval = FALSE,
  echo = TRUE,
  message = TRUE,
  warning = FALSE,
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction {.unnumbered}

**FIAwrapper** transforms raw USDA Forest Inventory and Analysis (FIA) database
tables into clean, imperial-unit tree lists using standard forestry terminology.
No more deciphering `CONDPROP_UNADJ`, `TPA_UNADJ`, or `SPCD` --- just `dbh`,
`height`, `tpa`, and `species`.

The package is designed for foresters, silviculturists, and forest biometricians
who need to:

- Build **per-acre tree lists** from FIA data for any area of interest
- Export tree lists directly to the **Forest Vegetation Simulator (FVS)**
- Analyze **remeasurement data** for growth modeling, mortality analysis, and
  stand dynamics
- Summarize **plot-level and species-level** attributes in familiar units
- Enrich plot data with **terrain, climate, and soils** from federal APIs
- Flag **outlier trees** using biologically plausible bounds
- Estimate **sampling error** for any attribute

Every function in FIAwrapper uses **imperial units** by default: DBH in inches,
height in feet, basal area in ft²/acre, and trees per acre (TPA). FIA database
column names like `DIA`, `HT`, `STATUSCD`, and `PLT_CN` are translated to
forestry terms like `dbh`, `height`, `status`, and `plot_cn`.

## How This Guide Is Organized

| Section | What You'll Learn |
|---------|-------------------|
| @sec-install | Installation, FVS variant setup, Python environment |
| @sec-fia-background | FIA program, plot design, table structure |
| @sec-data-acquisition | Downloading and loading FIA data |
| @sec-spatial | Filtering plots by shapefile or bounding box |
| @sec-tree-lists | Building clean tree lists from raw FIA data |
| @sec-plot-site | Plot summaries, site attributes, species composition |
| @sec-fvs | Exporting to FVS formats |
| @sec-time-series | Remeasurement, growth tracking, mortality |
| @sec-competition | Competition indices for growth modeling |
| @sec-quality | Outlier detection, variance estimation, regeneration |
| @sec-species | Species code system and lookups |
| @sec-utilities | Unit conversions, column name translation |
| @sec-spatial-apis | Terrain, climate, and soils from federal APIs |
| @sec-workflows | End-to-end workflows for common goals |

## Example Dataset

This guide uses `fia_ri`, a realistic FIA dataset bundled with the package.
It contains 12 inventory plots across 3 Rhode Island counties with 2
measurement cycles (2014 and 2019), representing typical southern New England
mixed forests. See @sec-example-data for details.

```{r}
library(FIAwrapper)
data(fia_ri)

names(fia_ri)
#> [1] "TREE" "PLOT" "COND"

nrow(fia_ri$TREE)
#> [1] 467

table(fia_ri$PLOT$INVYR)
#> 2014 2019
#>    6   12
```

# Installation & Setup {#sec-install}

## Installing the Package

```{r}
# Install from GitHub
devtools::install_github("ryanmismith/FIAwrapper")

library(FIAwrapper)
```

**Dependencies**: FIAwrapper imports `dplyr`, `tibble`, `rlang`, and `readr`.
Optional packages include `rFIA` (for data downloads), `sf` (for spatial
operations), `httr2` (for DataMart downloads), and `reticulate` (for Python
spatial backend).

## Setting Your FVS Variant {#sec-variant}

The FVS variant determines which species code mapping is used throughout the
session. Every function that works with species codes respects this setting.

```{r}
set_fvs_variant("NE")  # Northeast variant
#> FVS variant set to: NE
```

**Available variants** (see the [FVS website](https://www.fs.usda.gov/fvs/whatis.shtml)
for coverage maps):

| Code | Variant | Region |
|------|---------|--------|
| NE | Northeast | New England, Mid-Atlantic, Lake States |
| SN | Southern | Southeast US, Gulf states |
| LS | Lake States | WI, MI, MN |
| CS | Central States | Central hardwoods |
| PN | Pacific Northwest — Coast | Western OR, WA |
| WC | West Cascades | Cascade Range |
| BM | Blue Mountains | Eastern OR, WA |
| CR | Central Rockies | CO, eastern WY |
| EM | Eastern Montana | MT east of divide |
| IE | Inland Empire | Northern ID, western MT |
| SO | Southern Oregon | Southern OR |
| AK | Alaska | AK |

If you don't set a variant, FIAwrapper uses a default mapping table that covers
common species across regions. You can check the current variant at any time:

```{r}
getOption("FIAwrapper.variant")
#> [1] "NE"
```

::: {.callout-tip}
## When to Set a Variant
Set the variant **before** building tree lists or exporting to FVS. The variant
determines the 2-letter alpha code (e.g., "RM" for red maple in NE vs. "RM" in
SN) used in the `spp_code` column of your tree list.
:::

## Python Environment (Optional) {#sec-python-env}

Spatial functions (`get_plots_in_area()`, `get_terrain_data()`,
`get_climate_data()`, `get_soils_data()`) use Python libraries (geopandas,
shapely) behind the scenes via `reticulate`. The conda environment is created
automatically on first use.

```{r}
# Check if Python environment is ready
check_python_deps()
#> TRUE

# Manually create/update the environment
ensure_python_env()
```

If you don't need spatial functions, Python is never required. The core tree
list, summary, and FVS export functions are pure R.

::: {.callout-note}
## sf as Primary Spatial Engine
FIAwrapper uses the R `sf` package as the primary spatial engine when available.
Python (geopandas) is a fallback for environments where `sf` is hard to install.
Most users will never need the Python backend.
:::

# Understanding FIA Data {#sec-fia-background}

## The FIA Program

The USDA Forest Service [Forest Inventory and Analysis](https://www.fia.fs.usda.gov/)
(FIA) program is the nation's continuous forest census. FIA maintains a network
of ~320,000 permanent sample plots across all 50 states, remeasured on a
rotating 5-to-10-year cycle.

**Key references**:

- [FIADB User Guide](https://www.fia.fs.usda.gov/library/database-documentation/)
  --- Complete database documentation
- [FIA DataMart](https://apps.fs.usda.gov/fia/datamart/) --- Download raw data
- [EVALIDator](https://apps.fs.usda.gov/fiadb-api/evalidator) --- Web tool for
  pre-computed estimates
- Bechtold & Patterson (2005). *The Enhanced FIA Program — National Consistency
  with Regional Relevance*. Gen. Tech. Rep. SRS-80.
- Burrill et al. (2024). *The Forest Inventory and Analysis Database: Database
  Description and User Guide for Phase 2*. Version 9.2.

## FIA Plot Design {#sec-plot-design}

Each FIA plot is a **cluster of 4 subplots** arranged in a fixed pattern.
Understanding this design is critical for correct per-acre expansion.

```
     Subplot 2
        o
        |  120 ft
        |
Subplot 3 ---o--- Subplot 4
        |  120 ft (horizontal)
        |
     Subplot 1
     (plot center)
```

Each subplot contains two concentric sampling areas:

| Sample Area | Radius | Area | Trees Measured |
|------------|--------|------|----------------|
| **Subplot** | 24.0 ft (7.32 m) | 1,809.6 ft² | All trees ≥ 5.0 in DBH |
| **Microplot** | 6.8 ft (2.07 m) | 145.3 ft² | Saplings 1.0–4.9 in DBH |

**Per-acre expansion factors** (TPA) for a standard 4-subplot plot:

- Subplot trees (≥ 5.0 in DBH): `TPA = 1 / (π × 24² / 43560 × 4) ≈ 6.018`
- Microplot trees (< 5.0 in DBH): `TPA = 1 / (π × 6.8² / 43560 × 4) ≈ 74.965`

These expansion factors convert individual tree measurements to per-acre values.
FIAwrapper computes them automatically via `compute_tpa()` or within
`build_tree_list()`.

::: {.callout-important}
## Condition Proportion
When a plot straddles two forest conditions (e.g., half oak forest, half field),
FIA assigns a `CONDPROP_UNADJ` (0 to 1) to each condition. The TPA calculation
must account for this: `TPA_adjusted = TPA_base / CONDPROP_UNADJ`. FIAwrapper
handles this automatically when `CONDPROP_UNADJ` is available in the COND table.
:::

## FIA Database Tables {#sec-fia-tables}

FIA data is organized into relational tables. FIAwrapper uses three core tables:

### TREE Table
One row per tree per measurement. Key columns:

| FIA Column | FIAwrapper Name | Description |
|-----------|----------------|-------------|
| CN | tree_cn | Unique tree identifier |
| PLT_CN | plot_cn | Foreign key to PLOT table |
| SUBP | subplot | Subplot number (1–4) |
| STATUSCD | status_code | 1 = live, 2 = dead, 3 = removed |
| SPCD | spcd | FIA numeric species code (see [REF_SPECIES](https://apps.fs.usda.gov/fia/datamart/CSV/REF_SPECIES.csv)) |
| DIA | dbh | Diameter at breast height (inches) |
| HT | height | Total tree height (feet) |
| CR | crown_ratio | Compacted crown ratio (percent) |
| TPA_UNADJ | tpa_unadj | Unadjusted trees per acre |
| PREV_TRE_CN | prev_tree_cn | CN of same tree in previous measurement |
| VOLCFNET | vol_cf_net | Net cubic foot volume |
| VOLBFNET | vol_bf_net | Net board foot volume (Scribner) |
| DRYBIO_AG | biomass_ag_dry | Aboveground dry biomass (lbs) |
| CARBON_AG | carbon_ag | Aboveground carbon (lbs) |

### PLOT Table
One row per plot per measurement. Key columns:

| FIA Column | FIAwrapper Name | Description |
|-----------|----------------|-------------|
| CN | plot_cn | Unique plot identifier (use as join key) |
| STATECD | state_code | FIPS state code |
| COUNTYCD | county_code | FIPS county code |
| PLOT | plot_number | Plot number within county |
| INVYR | inventory_year | Year of measurement |
| LAT / LON | lat / lon | Approximate coordinates (fuzzed ±0.5–1 mi) |
| ELEV | elevation | Elevation (feet) |
| PREV_PLT_CN | — | CN of same plot in previous measurement |

::: {.callout-warning}
## Fuzzed Coordinates
FIA intentionally offsets plot coordinates by up to 1 mile for privacy. These
coordinates are adequate for regional analysis and spatial filtering, but should
**not** be used for precise site-level work. Exact coordinates require a
[confidential data agreement](https://www.fia.fs.usda.gov/tools-data/spatial/index.php).
:::

### COND Table
One row per condition per plot per measurement. Key columns:

| FIA Column | FIAwrapper Name | Description |
|-----------|----------------|-------------|
| PLT_CN | plot_cn | Foreign key to PLOT |
| CONDID | condition_id | Condition identifier within plot |
| CONDPROP_UNADJ | cond_proportion | Proportion of plot in this condition (0–1) |
| FORTYPCD | forest_type_code | Forest type code (see [REF_FOREST_TYPE](https://apps.fs.usda.gov/fia/datamart/CSV/REF_FOREST_TYPE.csv)) |
| STDAGE | stand_age | Stand age (years) |
| SICOND | site_index | Site index (base age 50, feet) |
| SITECLCD | site_class | Site productivity class (1–7) |
| SLOPE | slope | Slope in percent |
| ASPECT | aspect | Aspect in degrees (0–360) |
| BALIVE | ba_live | Live tree basal area (ft²/ac) on condition |
| OWNGRPCD | ownership_group | Ownership group code |

## How FIAwrapper Connects the Tables {#sec-table-joins}

When you call `build_tree_list()`, FIAwrapper performs these joins automatically:

1. **TREE → PLOT**: via `PLT_CN` to get location, year, and state info
2. **TREE → COND**: via `PLT_CN` + `CONDID` to get site attributes
3. **Species translation**: `SPCD` → `spp_code` (FVS alpha) and `species`
   (common name) using the active variant mapping
4. **TPA computation**: using `compute_tpa()` with the FIA subplot/microplot
   design
5. **Column renaming**: FIA jargon → forestry terms via `fia_to_forestry_names()`

The result is a single tibble with one row per tree and all measurements in
imperial units.

## The Example Dataset {#sec-example-data}

FIAwrapper ships with `fia_ri`, a realistic dataset modeled on Rhode Island
FIA data. It follows FIADB v9.2 specifications exactly and contains:

- **467 tree records** across 12 plots and 3 counties
- **2 inventory cycles** (2014, 2019) with 6 remeasured plots
- **15 species** typical of southern New England mixed forest
- Standing dead trees (~5% of stems) for mortality analysis
- Saplings on microplots for regeneration assessment

```{r}
data(fia_ri)

# Structure
str(fia_ri, max.level = 1)
#> List of 3
#>  $ TREE:'data.frame': 467 obs. of 22 variables
#>  $ PLOT:'data.frame': 18 obs. of 11 variables
#>  $ COND:'data.frame': 18 obs. of 16 variables

# Species in the dataset
sort(unique(fia_ri$TREE$SPCD))
#>  [1]  12  68  71  97 129 241 261 316 318 531 541 621 802 833 951

# What are these species?
set_fvs_variant("NE")
get_common_name(sort(unique(fia_ri$TREE$SPCD)))
```

This dataset is used in every example throughout this guide.
