---
title: "FIAwrapper: Complete Package Guide"
subtitle: "Clean FIA Tree Lists in Forestry Language"
author: "Ryan Smith"
date: last-modified
format:
  html:
    toc: true
    toc-depth: 4
    toc-expand: 2
    number-sections: true
    code-fold: false
    code-overflow: wrap
    theme: flatly
    highlight-style: github
    anchor-sections: true
    smooth-scroll: true
    link-external-newwindow: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  eval = FALSE,
  echo = TRUE,
  message = TRUE,
  warning = FALSE,
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction {.unnumbered}

**FIAwrapper** transforms raw USDA Forest Inventory and Analysis (FIA) database
tables into clean, imperial-unit tree lists using standard forestry terminology.
No more deciphering `CONDPROP_UNADJ`, `TPA_UNADJ`, or `SPCD` --- just `dbh`,
`height`, `tpa`, and `species`.

The package is designed for foresters, silviculturists, and forest biometricians
who need to:

- Build **per-acre tree lists** from FIA data for any area of interest
- Export tree lists directly to the **Forest Vegetation Simulator (FVS)**
- Analyze **remeasurement data** for growth modeling, mortality analysis, and
  stand dynamics
- Summarize **plot-level and species-level** attributes in familiar units
- Enrich plot data with **terrain, climate, and soils** from federal APIs
- Flag **outlier trees** using biologically plausible bounds
- Estimate **sampling error** for any attribute

Every function in FIAwrapper uses **imperial units** by default: DBH in inches,
height in feet, basal area in ft²/acre, and trees per acre (TPA). FIA database
column names like `DIA`, `HT`, `STATUSCD`, and `PLT_CN` are translated to
forestry terms like `dbh`, `height`, `status`, and `plot_cn`.

## How This Guide Is Organized

| Section | What You'll Learn |
|---------|-------------------|
| @sec-install | Installation, FVS variant setup, Python environment |
| @sec-fia-background | FIA program, plot design, table structure |
| @sec-data-acquisition | Downloading and loading FIA data |
| @sec-spatial | Filtering plots by shapefile or bounding box |
| @sec-tree-lists | Building clean tree lists from raw FIA data |
| @sec-plot-site | Plot summaries, site attributes, species composition |
| @sec-fvs | Exporting to FVS formats |
| @sec-time-series | Remeasurement, growth tracking, mortality |
| @sec-competition | Competition indices for growth modeling |
| @sec-quality | Outlier detection, variance estimation, regeneration |
| @sec-species | Species code system and lookups |
| @sec-utilities | Unit conversions, column name translation |
| @sec-spatial-apis | Terrain, climate, and soils from federal APIs |
| @sec-workflows | End-to-end workflows for common goals |

## Example Dataset

This guide uses `fia_ri`, a realistic FIA dataset bundled with the package.
It contains 12 inventory plots across 3 Rhode Island counties with 2
measurement cycles (2014 and 2019), representing typical southern New England
mixed forests. See @sec-example-data for details.

```{r}
library(FIAwrapper)
data(fia_ri)

names(fia_ri)
#> [1] "TREE" "PLOT" "COND"

nrow(fia_ri$TREE)
#> [1] 467

table(fia_ri$PLOT$INVYR)
#> 2014 2019
#>    6   12
```

# Installation & Setup {#sec-install}

## Installing the Package

```{r}
# Install from GitHub
devtools::install_github("ryanmismith/FIAwrapper")

library(FIAwrapper)
```

**Dependencies**: FIAwrapper imports `dplyr`, `tibble`, `rlang`, and `readr`.
Optional packages include `rFIA` (for data downloads), `sf` (for spatial
operations), `httr2` (for DataMart downloads), and `reticulate` (for Python
spatial backend).

## Setting Your FVS Variant {#sec-variant}

The FVS variant determines which species code mapping is used throughout the
session. Every function that works with species codes respects this setting.

```{r}
set_fvs_variant("NE")  # Northeast variant
#> FVS variant set to: NE
```

**Available variants** (see the [FVS website](https://www.fs.usda.gov/fvs/whatis.shtml)
for coverage maps):

| Code | Variant | Region |
|------|---------|--------|
| NE | Northeast | New England, Mid-Atlantic, Lake States |
| SN | Southern | Southeast US, Gulf states |
| LS | Lake States | WI, MI, MN |
| CS | Central States | Central hardwoods |
| PN | Pacific Northwest — Coast | Western OR, WA |
| WC | West Cascades | Cascade Range |
| BM | Blue Mountains | Eastern OR, WA |
| CR | Central Rockies | CO, eastern WY |
| EM | Eastern Montana | MT east of divide |
| IE | Inland Empire | Northern ID, western MT |
| SO | Southern Oregon | Southern OR |
| AK | Alaska | AK |

If you don't set a variant, FIAwrapper uses a default mapping table that covers
common species across regions. You can check the current variant at any time:

```{r}
getOption("FIAwrapper.variant")
#> [1] "NE"
```

::: {.callout-tip}
## When to Set a Variant
Set the variant **before** building tree lists or exporting to FVS. The variant
determines the 2-letter alpha code (e.g., "RM" for red maple in NE vs. "RM" in
SN) used in the `spp_code` column of your tree list.
:::

## Python Environment (Optional) {#sec-python-env}

Spatial functions (`get_plots_in_area()`, `get_terrain_data()`,
`get_climate_data()`, `get_soils_data()`) use Python libraries (geopandas,
shapely) behind the scenes via `reticulate`. The conda environment is created
automatically on first use.

```{r}
# Check if Python environment is ready
check_python_deps()
#> TRUE

# Manually create/update the environment
ensure_python_env()
```

If you don't need spatial functions, Python is never required. The core tree
list, summary, and FVS export functions are pure R.

::: {.callout-note}
## sf as Primary Spatial Engine
FIAwrapper uses the R `sf` package as the primary spatial engine when available.
Python (geopandas) is a fallback for environments where `sf` is hard to install.
Most users will never need the Python backend.
:::

# Understanding FIA Data {#sec-fia-background}

## The FIA Program

The USDA Forest Service [Forest Inventory and Analysis](https://www.fia.fs.usda.gov/)
(FIA) program is the nation's continuous forest census. FIA maintains a network
of ~320,000 permanent sample plots across all 50 states, remeasured on a
rotating 5-to-10-year cycle.

**Key references**:

- [FIADB User Guide](https://www.fia.fs.usda.gov/library/database-documentation/)
  --- Complete database documentation
- [FIA DataMart](https://apps.fs.usda.gov/fia/datamart/) --- Download raw data
- [EVALIDator](https://apps.fs.usda.gov/fiadb-api/evalidator) --- Web tool for
  pre-computed estimates
- Bechtold & Patterson (2005). *The Enhanced FIA Program — National Consistency
  with Regional Relevance*. Gen. Tech. Rep. SRS-80.
- Burrill et al. (2024). *The Forest Inventory and Analysis Database: Database
  Description and User Guide for Phase 2*. Version 9.2.

## FIA Plot Design {#sec-plot-design}

Each FIA plot is a **cluster of 4 subplots** arranged in a fixed pattern.
Understanding this design is critical for correct per-acre expansion.

```
     Subplot 2
        o
        |  120 ft
        |
Subplot 3 ---o--- Subplot 4
        |  120 ft (horizontal)
        |
     Subplot 1
     (plot center)
```

Each subplot contains two concentric sampling areas:

| Sample Area | Radius | Area | Trees Measured |
|------------|--------|------|----------------|
| **Subplot** | 24.0 ft (7.32 m) | 1,809.6 ft² | All trees ≥ 5.0 in DBH |
| **Microplot** | 6.8 ft (2.07 m) | 145.3 ft² | Saplings 1.0–4.9 in DBH |

**Per-acre expansion factors** (TPA) for a standard 4-subplot plot:

- Subplot trees (≥ 5.0 in DBH): `TPA = 1 / (π × 24² / 43560 × 4) ≈ 6.018`
- Microplot trees (< 5.0 in DBH): `TPA = 1 / (π × 6.8² / 43560 × 4) ≈ 74.965`

These expansion factors convert individual tree measurements to per-acre values.
FIAwrapper computes them automatically via `compute_tpa()` or within
`build_tree_list()`.

::: {.callout-important}
## Condition Proportion
When a plot straddles two forest conditions (e.g., half oak forest, half field),
FIA assigns a `CONDPROP_UNADJ` (0 to 1) to each condition. The TPA calculation
must account for this: `TPA_adjusted = TPA_base / CONDPROP_UNADJ`. FIAwrapper
handles this automatically when `CONDPROP_UNADJ` is available in the COND table.
:::

## FIA Database Tables {#sec-fia-tables}

FIA data is organized into relational tables. FIAwrapper uses three core tables:

### TREE Table
One row per tree per measurement. Key columns:

| FIA Column | FIAwrapper Name | Description |
|-----------|----------------|-------------|
| CN | tree_cn | Unique tree identifier |
| PLT_CN | plot_cn | Foreign key to PLOT table |
| SUBP | subplot | Subplot number (1–4) |
| STATUSCD | status_code | 1 = live, 2 = dead, 3 = removed |
| SPCD | spcd | FIA numeric species code (see [REF_SPECIES](https://apps.fs.usda.gov/fia/datamart/CSV/REF_SPECIES.csv)) |
| DIA | dbh | Diameter at breast height (inches) |
| HT | height | Total tree height (feet) |
| CR | crown_ratio | Compacted crown ratio (percent) |
| TPA_UNADJ | tpa_unadj | Unadjusted trees per acre |
| PREV_TRE_CN | prev_tree_cn | CN of same tree in previous measurement |
| VOLCFNET | vol_cf_net | Net cubic foot volume |
| VOLBFNET | vol_bf_net | Net board foot volume (Scribner) |
| DRYBIO_AG | biomass_ag_dry | Aboveground dry biomass (lbs) |
| CARBON_AG | carbon_ag | Aboveground carbon (lbs) |

### PLOT Table
One row per plot per measurement. Key columns:

| FIA Column | FIAwrapper Name | Description |
|-----------|----------------|-------------|
| CN | plot_cn | Unique plot identifier (use as join key) |
| STATECD | state_code | FIPS state code |
| COUNTYCD | county_code | FIPS county code |
| PLOT | plot_number | Plot number within county |
| INVYR | inventory_year | Year of measurement |
| LAT / LON | lat / lon | Approximate coordinates (fuzzed ±0.5–1 mi) |
| ELEV | elevation | Elevation (feet) |
| PREV_PLT_CN | — | CN of same plot in previous measurement |

::: {.callout-warning}
## Fuzzed Coordinates
FIA intentionally offsets plot coordinates by up to 1 mile for privacy. These
coordinates are adequate for regional analysis and spatial filtering, but should
**not** be used for precise site-level work. Exact coordinates require a
[confidential data agreement](https://www.fia.fs.usda.gov/tools-data/spatial/index.php).
:::

### COND Table
One row per condition per plot per measurement. Key columns:

| FIA Column | FIAwrapper Name | Description |
|-----------|----------------|-------------|
| PLT_CN | plot_cn | Foreign key to PLOT |
| CONDID | condition_id | Condition identifier within plot |
| CONDPROP_UNADJ | cond_proportion | Proportion of plot in this condition (0–1) |
| FORTYPCD | forest_type_code | Forest type code (see [REF_FOREST_TYPE](https://apps.fs.usda.gov/fia/datamart/CSV/REF_FOREST_TYPE.csv)) |
| STDAGE | stand_age | Stand age (years) |
| SICOND | site_index | Site index (base age 50, feet) |
| SITECLCD | site_class | Site productivity class (1–7) |
| SLOPE | slope | Slope in percent |
| ASPECT | aspect | Aspect in degrees (0–360) |
| BALIVE | ba_live | Live tree basal area (ft²/ac) on condition |
| OWNGRPCD | ownership_group | Ownership group code |

## How FIAwrapper Connects the Tables {#sec-table-joins}

When you call `build_tree_list()`, FIAwrapper performs these joins automatically:

1. **TREE → PLOT**: via `PLT_CN` to get location, year, and state info
2. **TREE → COND**: via `PLT_CN` + `CONDID` to get site attributes
3. **Species translation**: `SPCD` → `spp_code` (FVS alpha) and `species`
   (common name) using the active variant mapping
4. **TPA computation**: using `compute_tpa()` with the FIA subplot/microplot
   design
5. **Column renaming**: FIA jargon → forestry terms via `fia_to_forestry_names()`

The result is a single tibble with one row per tree and all measurements in
imperial units.

## The Example Dataset {#sec-example-data}

FIAwrapper ships with `fia_ri`, a realistic dataset modeled on Rhode Island
FIA data. It follows FIADB v9.2 specifications exactly and contains:

- **467 tree records** across 12 plots and 3 counties
- **2 inventory cycles** (2014, 2019) with 6 remeasured plots
- **15 species** typical of southern New England mixed forest
- Standing dead trees (~5% of stems) for mortality analysis
- Saplings on microplots for regeneration assessment

```{r}
data(fia_ri)

# Structure
str(fia_ri, max.level = 1)
#> List of 3
#>  $ TREE:'data.frame': 467 obs. of 22 variables
#>  $ PLOT:'data.frame': 18 obs. of 11 variables
#>  $ COND:'data.frame': 18 obs. of 16 variables

# Species in the dataset
sort(unique(fia_ri$TREE$SPCD))
#>  [1]  12  68  71  97 129 241 261 316 318 531 541 621 802 833 951

# What are these species?
set_fvs_variant("NE")
get_common_name(sort(unique(fia_ri$TREE$SPCD)))
```

This dataset is used in every example throughout this guide.

# Data Acquisition {#sec-data-acquisition}

FIAwrapper provides two functions for getting FIA data onto your machine:
`fetch_fia_data()` for downloading, and `load_fia_data()` for loading
previously downloaded data.

## Downloading FIA Data

`fetch_fia_data()` downloads FIA tables for one or more states. It tries the
[rFIA](https://cran.r-project.org/package=rFIA) package first (if installed),
then falls back to direct CSV downloads from the
[FIA DataMart](https://apps.fs.usda.gov/fia/datamart/).

```{r}
# Download core tables for Maine and New Hampshire
fia <- fetch_fia_data(
  states = c("ME", "NH"),
  tables = c("TREE", "PLOT", "COND", "SUBPLOT"),
  data_dir = "~/fia_data",
  method = "auto"  # tries rFIA first, then DataMart
)

names(fia)
#> [1] "TREE" "PLOT" "COND" "SUBPLOT"
```

**Parameters**:

- `states`: Character vector of 2-letter state abbreviations (e.g., `c("ME", "NH", "VT")`)
- `tables`: Which FIA tables to download. Default: `c("TREE", "PLOT", "COND", "SUBPLOT")`
- `data_dir`: Directory to cache downloads. Default: `getOption("FIAwrapper.data_dir")` or `tempdir()`
- `method`: `"auto"` (default), `"rfia"`, or `"datamart"`

**Caching**: Downloaded files are cached locally. Files less than 30 days old are
reused automatically. Delete the cache directory to force a fresh download.

::: {.callout-tip}
## rFIA vs DataMart
The `rFIA` method is faster and more reliable for multi-state downloads. Install
it with `install.packages("rFIA")`. The DataMart fallback downloads individual
CSV files from [apps.fs.usda.gov/fia/datamart/CSV/](https://apps.fs.usda.gov/fia/datamart/CSV/).
:::

## Loading Cached Data

```{r}
# Load previously downloaded data
fia <- load_fia_data("~/fia_data", states = "ME")
```

`load_fia_data()` auto-detects whether data was saved in rFIA format (RDS files)
or as raw CSVs. Optionally filter by state when loading multi-state datasets.

## Using the Example Dataset

For this guide, we use the bundled `fia_ri` dataset:

```{r}
data(fia_ri)
# This is equivalent to what fetch_fia_data() returns:
# a named list with TREE, PLOT, and COND data frames
```

# Spatial Queries {#sec-spatial}

FIAwrapper can filter FIA plots to a specific area of interest (AOI) using
shapefiles, `sf` objects, or bounding boxes.

## Filtering Plots by Area

`get_plots_in_area()` is the core spatial function. It finds all FIA plots
within your AOI and returns a filtered copy of the FIA data.

```{r}
# From a shapefile
fia_local <- get_plots_in_area(
  aoi = "my_property.shp",
  fia_data = fia,
  buffer_ft = 5280  # 1-mile buffer around the AOI
)

# From an sf object
library(sf)
my_bbox <- st_as_sf(st_as_sfc(st_bbox(
  c(xmin = -71.6, ymin = 41.4, xmax = -71.3, ymax = 41.9),
  crs = 4326
)))
fia_local <- get_plots_in_area(aoi = my_bbox, fia_data = fia_ri)

# From a bounding box vector (xmin, ymin, xmax, ymax)
fia_local <- get_plots_in_area(
  aoi = c(-71.6, 41.4, -71.3, 41.9),
  fia_data = fia_ri
)
```

**Parameters**:

- `aoi`: File path (shapefile, GeoPackage, GeoJSON), `sf` object, or numeric
  vector `c(xmin, ymin, xmax, ymax)` in WGS84
- `fia_data`: FIA data list (from `fetch_fia_data()`) or character vector of
  state abbreviations (auto-downloads)
- `buffer_ft`: Buffer distance in feet around the AOI (default: 0)
- `data_dir`: Cache directory when `fia_data` is state abbreviations

**Returns**: A filtered FIA data list (same structure as input) containing only
plots within the AOI. All tables (TREE, PLOT, COND) are filtered consistently.

::: {.callout-note}
## Spatial Engine
`get_plots_in_area()` uses the R `sf` package when available. If `sf` is not
installed, it falls back to Python geopandas via `reticulate`. Install `sf`
with `install.packages("sf")` for the simplest setup.
:::

## Extracting Plot Coordinates

`plot_locations()` creates an `sf` POINT dataset from FIA plot coordinates.

```{r}
plots_sf <- plot_locations(fia_ri)
class(plots_sf)
#> [1] "sf"         "data.frame"

head(plots_sf[, c("plot_cn", "inventory_year")])
```

**Returns**: An `sf` data frame with POINT geometry in the specified CRS
(default: EPSG:4326 / WGS84). Includes all PLOT table columns.

## Mapping Plots

`map_plots()` creates a quick interactive or static map.

```{r}
# Basic map
map_plots(fia_ri)

# With AOI overlay
map_plots(fia_ri, aoi = "my_property.shp")
```

Uses `leaflet` for interactive maps (if installed), otherwise falls back to
base R `plot()`. The AOI boundary is overlaid when provided.

# Building Tree Lists {#sec-tree-lists}

## `build_tree_list()` — The Core Function {#sec-build-tree-list}

This is the main workhorse function. It transforms raw FIA tables into a clean,
analysis-ready tree list.

```{r}
set_fvs_variant("NE")

trees <- build_tree_list(
  fia_data = fia_ri,
  include_dead = FALSE,
  variant = "NE",
  most_recent = TRUE,
  compute_expansion = TRUE
)

nrow(trees)
#> [1] 452

head(trees[, c("plot_id", "spp_code", "species", "dbh", "height", "tpa", "ba_per_acre")])
```

**Parameters**:

- `fia_data`: Named list with TREE, PLOT, and COND data frames
- `include_dead`: Include standing dead trees? Default `FALSE`. When `TRUE`,
  adds a `status` column ("live" / "dead")
- `variant`: FVS variant code for species translation. Default: session variant
- `most_recent`: Keep only the latest inventory cycle? Default `TRUE`
- `compute_expansion`: Compute TPA from subplot geometry? Default `TRUE`.
  When `FALSE`, uses FIA's `TPA_UNADJ` directly

### Output Columns {#sec-tree-list-columns}

The output tibble contains one row per tree with these columns:

**Tree identity and location**:

| Column | Type | Description |
|--------|------|-------------|
| `plot_id` | character | Unique plot ID: `STATECD_COUNTYCD_PLOT` |
| `plot_cn` | integer | FIA plot control number |
| `subplot` | integer | Subplot number (1–4) |
| `tree_num` | integer | Tree number within subplot |
| `tree_cn` | integer | Unique tree control number |
| `inventory_year` | integer | Year of measurement |

**Species**:

| Column | Type | Description |
|--------|------|-------------|
| `species` | character | Common name (e.g., "red maple") |
| `spp_code` | character | FVS alpha code (e.g., "RM") |
| `spcd` | integer | FIA numeric code (e.g., 316), kept for reference |

**Measurements** (imperial units):

| Column | Type | Description |
|--------|------|-------------|
| `dbh` | numeric | Diameter at breast height (inches) |
| `height` | numeric | Total height (feet) |
| `crown_ratio` | numeric | Crown ratio (percent, 0–100) |
| `crown_class` | integer | Crown class (1=open-grown, 2=dominant, 3=codominant, 4=intermediate, 5=overtopped) |

**Per-acre values**:

| Column | Type | Description |
|--------|------|-------------|
| `tpa` | numeric | Trees per acre this tree represents |
| `ba_tree` | numeric | Individual tree basal area (ft²) |
| `ba_per_acre` | numeric | BA contribution = `ba_tree × tpa` (ft²/acre) |

**Status**:

| Column | Type | Description |
|--------|------|-------------|
| `status` | character | "live" or "dead" |
| `status_code` | integer | FIA STATUSCD (1, 2, or 3) |
| `decay_class` | integer | Decay class for dead trees (1–5) |
| `damage_code1` | integer | Primary damage agent code |

**Volume and biomass**:

| Column | Type | Description |
|--------|------|-------------|
| `vol_cf_net` | numeric | Net cubic foot volume |
| `vol_cf_gross` | numeric | Gross cubic foot volume |
| `vol_bf_net` | numeric | Net board foot volume (Scribner) |
| `biomass_ag_dry` | numeric | Aboveground dry biomass (lbs) |
| `carbon_ag` | numeric | Aboveground carbon (lbs) |

**Site and location**:

| Column | Type | Description |
|--------|------|-------------|
| `state_code` | integer | FIPS state code |
| `county_code` | integer | FIPS county code |
| `lat` / `lon` | numeric | Fuzzed coordinates |
| `elevation` | numeric | Elevation (feet) |
| `slope` | numeric | Slope (percent) |
| `aspect` | numeric | Aspect (degrees) |
| `site_index` | numeric | Site index (base age 50, feet) |
| `site_class` | integer | Site productivity class |
| `stand_age` | integer | Stand age (years) |
| `forest_type_code` | integer | FIA forest type code |

### How TPA Is Computed {#sec-tpa-computation}

`build_tree_list()` computes TPA using the FIA nested plot design:

1. Trees with **DBH ≥ 5.0 inches** are measured on the **subplot** (24 ft
   radius). Each subplot covers `π × 24² / 43560 = 0.04154 acres`. With 4
   subplots, each tree represents `1 / (0.04154 × 4) ≈ 6.018 trees/acre`.

2. Trees with **DBH < 5.0 inches** are measured on the **microplot** (6.8 ft
   radius). Each microplot covers `π × 6.8² / 43560 = 0.003336 acres`. With
   4 microplots, each tree represents `1 / (0.003336 × 4) ≈ 74.965 trees/acre`.

3. If `CONDPROP_UNADJ` < 1.0 (plot straddles conditions), TPA is adjusted:
   `TPA = TPA_base / CONDPROP_UNADJ`.

::: {.callout-important}
## TPA vs TPA_UNADJ
FIA provides `TPA_UNADJ` in the TREE table, which is their own expansion
factor. FIAwrapper's `compute_tpa()` recalculates this from first principles
using the subplot geometry. In most cases the values agree closely. Set
`compute_expansion = FALSE` to use FIA's values instead.
:::

### Working with Dead Trees

```{r}
all_trees <- build_tree_list(fia_ri, include_dead = TRUE)

# Status breakdown
table(all_trees$status)
#>  dead  live
#>    15   452

# Dead tree attributes
snags <- all_trees[all_trees$status == "dead", ]
head(snags[, c("spp_code", "dbh", "height", "decay_class", "tpa")])
```

Dead trees (STATUSCD = 2) are standing dead / snags. They have valid DBH and
height measurements, plus a `decay_class` (1 = recently dead, 5 = heavily
decomposed). Mortality analysis requires `include_dead = TRUE`.

## `get_tree_list()` — One-Line Convenience

`get_tree_list()` chains `fetch_fia_data()` → `get_plots_in_area()` →
`build_tree_list()` into a single call.

```{r}
# Download + clip + build in one step
trees <- get_tree_list(
  states = "ME",
  aoi = "my_property.shp",
  data_dir = "~/fia_data",
  include_dead = TRUE,
  variant = "NE",
  most_recent = TRUE
)
```

**Parameters**: Combines all parameters from the three underlying functions.
Provide `states` and/or `aoi` to control data scope.

## Low-Level Data Prep: `prep_fia_data()`

For more control over the table joining process, use `prep_fia_data()` directly.
This function joins TREE, PLOT, and COND tables and renames columns, but does
**not** compute TPA or filter by status.

```{r}
prepped <- prep_fia_data(
  fia_trees = fia_ri$TREE,
  fia_plots = fia_ri$PLOT,
  fia_cond = fia_ri$COND,
  include_dead = TRUE,
  variant = "NE"
)

# All FIA columns renamed to forestry terms
names(prepped)
```

Useful when you want to apply custom filtering or compute expansion factors
differently (e.g., for non-standard plot designs).

# Plot & Site Data {#sec-plot-site}

## Plot Summaries {#sec-plot-summary}

`get_plot_summary()` aggregates a tree list to one row per plot with standard
forestry metrics.

```{r}
trees <- build_tree_list(fia_ri, variant = "NE")
plots <- get_plot_summary(trees)

head(plots[, c("plot_id", "tpa", "ba_per_acre", "qmd", "top_height",
               "species_count", "dominant_species")])
```

**Output columns**:

| Column | Units | Description |
|--------|-------|-------------|
| `tpa` | trees/acre | Total trees per acre |
| `ba_per_acre` | ft²/acre | Total basal area per acre |
| `qmd` | inches | Quadratic mean diameter |
| `mean_dbh` | inches | TPA-weighted mean DBH |
| `top_height` | feet | Mean height of tallest 40 TPA trees (site height proxy) |
| `mean_height` | feet | TPA-weighted mean height |
| `species_count` | count | Number of species on plot |
| `dominant_species` | character | Species with highest BA |
| `vol_per_acre` | ft³/acre | Net cubic foot volume (if available) |
| `n_trees` | count | Number of measured trees |

**Dead tree handling**: By default, `get_plot_summary()` excludes dead trees.
Set `include_dead = TRUE` to include them in all calculations.

```{r}
# Compare live-only vs all-stems summaries
plots_live <- get_plot_summary(all_trees, include_dead = FALSE)
plots_all  <- get_plot_summary(all_trees, include_dead = TRUE)

# Dead trees add to TPA and BA
mean(plots_all$tpa - plots_live$tpa)
```

::: {.callout-tip}
## QMD Computation
QMD (quadratic mean diameter) is computed as:
$$QMD = \sqrt{\frac{\sum_{i=1}^{n} DBH_i^2 \times TPA_i}{\sum_{i=1}^{n} TPA_i}}$$
This is the diameter of the tree with average basal area, and is always
larger than the arithmetic mean diameter. It's the standard metric used in
silvicultural prescriptions and yield tables.
:::

## Site Attributes {#sec-site-data}

`get_site_data()` extracts plot-level site and stand characteristics from the
PLOT and COND tables.

```{r}
site <- get_site_data(fia_ri)

head(site[, c("plot_cn", "inventory_year", "elevation", "slope", "aspect",
              "stand_age", "site_index", "forest_type_code")])
```

**Output columns**:

| Column | Units | Description |
|--------|-------|-------------|
| `elevation` | feet | Elevation above sea level |
| `slope` | percent | Slope gradient |
| `aspect` | degrees | Aspect (0 = north, 90 = east, 180 = south, 270 = west) |
| `forest_type_code` | integer | FIA forest type (see [REF_FOREST_TYPE](https://apps.fs.usda.gov/fia/datamart/CSV/REF_FOREST_TYPE.csv)) |
| `stand_age` | years | Stand age |
| `site_index` | feet | Site index at base age 50 (height of dominant trees) |
| `site_class` | integer | Site productivity class (1=highest, 7=lowest) |
| `ownership_group` | integer | Ownership code ([ref](https://www.fia.fs.usda.gov/library/database-documentation/)) |
| `stand_density_index` | index | Reineke's stand density index |
| `ba_live` | ft²/acre | Live basal area on condition |

**Merging with plot summaries**:

```{r}
# Combine metrics with site data
full_plots <- merge(plots, site, by = "plot_cn", suffixes = c("", "_site"))

# Now you have BA/acre + site index + slope + etc. in one table
head(full_plots[, c("plot_id", "ba_per_acre", "qmd", "site_index", "slope", "stand_age")])
```

## Species Composition {#sec-species-summary}

`get_species_summary()` creates per-species summaries for each plot.

```{r}
spp <- get_species_summary(trees)

head(spp[, c("plot_id", "spp_code", "species_name", "tpa", "ba_per_acre", "qmd", "pct_ba")])
```

**Output columns**:

| Column | Description |
|--------|-------------|
| `spp_code` | FVS alpha species code |
| `species_name` | Common name |
| `tpa` | Trees per acre for this species |
| `ba_per_acre` | Basal area (ft²/acre) for this species |
| `qmd` | QMD for this species (inches) |
| `mean_dbh` | Weighted mean DBH (inches) |
| `pct_ba` | Percent of total plot BA from this species |
| `n_trees` | Number of measured trees |

**Finding dominant species across all plots**:

```{r}
# Top species by mean BA/acre across all plots
spp |>
  dplyr::group_by(spp_code, species_name) |>
  dplyr::summarize(
    mean_ba = mean(ba_per_acre, na.rm = TRUE),
    plots_present = dplyr::n(),
    .groups = "drop"
  ) |>
  dplyr::arrange(dplyr::desc(mean_ba)) |>
  head(10)
```

### Relationship Between Summary Functions

```
build_tree_list()         # Tree-level data (1 row per tree)
    │
    ├── get_plot_summary()    # Plot-level (1 row per plot)
    │       Aggregates: TPA, BA/ac, QMD, species count
    │
    ├── get_species_summary() # Species×Plot level
    │       Aggregates: TPA, BA/ac, QMD per species per plot
    │
    └── get_site_data()       # Plot-level site attributes
            Source: PLOT + COND tables directly
```

## `summarize_plot()` — Vector-Based Alternative

For users who prefer working with raw vectors (not tibbles), `summarize_plot()`
accepts stand, plot, DBH, height, and TPA as separate vectors.

```{r}
result <- summarize_plot(
  stand   = trees$plot_id,
  plot    = trees$plot_cn,
  dbh     = trees$dbh,
  height  = trees$height,
  tpa     = trees$tpa,
  species = trees$spp_code
)
head(result)
```

This function also supports custom aggregation columns via `sum_cols` and
`mean_cols` parameters.
