---
title: "FIAwrapper: Clean FIA Tree Lists in Forestry Language"
author: "Ryan Smith"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  eval = FALSE,
  echo = TRUE,
  message = TRUE,
  warning = FALSE
)
```

# Introduction

**FIAwrapper** transforms USDA Forest Inventory and Analysis (FIA) data into
clean, imperial-unit tree lists using standard forestry terminology. No more
deciphering `CONDPROP_UNADJ`, `TPA_UNADJ`, or `SPCD` codes --- just
`dbh`, `height`, `tpa`, and `species`.

Key capabilities:

- **Shapefile → tree list** in one line of code
- **Imperial units** throughout (DBH in inches, height in feet, BA in ft²/acre)
- **FVS-ready exports** for any variant
- **Remeasurement time series** with growth increments and mortality tracking
- **Competition indices** (BAL, relative DBH) ready for growth modeling
- **Spatial enrichment** with terrain, climate, and soils data from federal APIs
- **Python spatial backend** managed automatically --- you only need R

# 1. Installation & Setup

```{r install}
# Install from GitHub
devtools::install_github("ryanmismith/FIAwrapper")

library(FIAwrapper)
```

## Set Your FVS Variant

This determines species code mappings throughout your session.

```{r variant}
set_fvs_variant("NE")  # Northeast variant
# Other options: "SN" (Southern), "LS" (Lake States), "CS", "PN", etc.
```

## Python Environment (Automatic)

Spatial functions use Python (geopandas, shapely) behind the scenes.
The conda environment is created automatically on first use:

```{r python-check}
# Check status (optional)
check_python_deps()
```

# 2. Quick Start

## One-Line Tree List

```{r quickstart}
# Get a tree list for Rhode Island
trees <- get_tree_list(states = "RI")
head(trees)
```

## From a Shapefile

```{r quickstart-shp}
# Shapefile → tree list
trees <- get_tree_list(
  states = "ME",
  aoi = "my_property.shp",
  include_dead = TRUE
)
```

# 3. Working with FIA Data

## Downloading Data

```{r fetch}
# Download via rFIA (if installed) or FIA DataMart
fia <- fetch_fia_data(
  states = c("ME", "NH"),
  data_dir = "~/fia_data"
)

# What tables did we get?
names(fia)
# [1] "TREE" "PLOT" "COND" "SUBPLOT"

nrow(fia$TREE)
nrow(fia$PLOT)
```

## Loading Cached Data

```{r load}
# Reuse previously downloaded data
fia <- load_fia_data("~/fia_data", states = "ME")
```

# 4. Spatial Queries

## Filter Plots by Area

```{r spatial}
# From a shapefile
fia_local <- get_plots_in_area(
  aoi = "my_property.shp",
  fia_data = fia,
  buffer_ft = 5280  # 1-mile buffer
)

# From an sf object
library(sf)
bbox <- st_as_sf(st_as_sfc(st_bbox(
  c(xmin = -70, ymin = 44, xmax = -69, ymax = 45),
  crs = 4326
)))
fia_local <- get_plots_in_area(aoi = bbox, fia_data = fia)
```

## Map Plot Locations

```{r map}
# Interactive map (leaflet) or static plot
map_plots(fia_local, aoi = "my_property.shp")
```

## Extract Plot Coordinates

```{r coords}
plots_sf <- plot_locations(fia)
# Note: FIA coordinates are fuzzed (~0.5-1 mile offset)
```

# 5. Building Tree Lists

## The Full Pipeline

```{r treelist}
trees <- build_tree_list(
  fia_data = fia_local,
  include_dead = FALSE,    # Live trees only
  variant = "NE",          # NE FVS species codes
  most_recent = TRUE       # Latest inventory cycle only
)

# What columns do we get?
names(trees)
# plot_id, subplot, tree_num, inventory_year,
# species, spp_code, spcd, dbh, height, crown_ratio,
# tpa, ba_tree, ba_per_acre, status, ...
```

## Understanding the Output

| Column | Description | Units |
|--------|------------|-------|
| `dbh` | Diameter at breast height | inches |
| `height` | Total height | feet |
| `tpa` | Trees per acre (expansion factor) | trees/acre |
| `ba_tree` | Individual tree basal area | ft² |
| `ba_per_acre` | BA contribution = ba_tree × tpa | ft²/acre |
| `species` | Common name | text |
| `spp_code` | FVS alpha code | 2-letter |
| `crown_ratio` | Crown ratio | percent |

## Include Dead Trees

```{r dead}
all_trees <- build_tree_list(fia_local, include_dead = TRUE)
table(all_trees$status)
#  dead  live
#   342  4567

# Dead tree attributes
snags <- all_trees[all_trees$status == "dead", ]
table(snags$decay_class)
```

# 6. Plot & Site Data

## Plot Summaries

```{r plot-summary}
plots <- get_plot_summary(trees)
head(plots)
# plot_id, tpa, ba_per_acre, qmd, mean_dbh, top_height,
# species_count, dominant_species, ...

# Average BA/acre across plots
mean(plots$ba_per_acre)
```

## Site Attributes

```{r site}
site <- get_site_data(fia_local)
head(site)
# plot_cn, elevation, slope, aspect, forest_type_code,
# stand_age, site_index, ownership_code, ...
```

## Species Composition

```{r species}
spp <- get_species_summary(trees)
head(spp)
# plot_id, spp_code, species_name, tpa, ba_per_acre, qmd, pct_ba

# Top species by BA
spp |>
  dplyr::group_by(spp_code, species_name) |>
  dplyr::summarize(mean_ba = mean(ba_per_acre)) |>
  dplyr::arrange(dplyr::desc(mean_ba)) |>
  head(10)
```

# 7. FVS Export

## Tree List for FVS

```{r fvs-tree}
# Fixed-width FVS format
export_fvs_treelist(trees, "fvs_input.tre", variant = "NE")

# CSV format
export_fvs_treelist(trees, "fvs_input.csv", format = "csv")

# Suppose-compatible format
export_fvs_treelist(trees, "fvs_suppose.csv", format = "suppose")
```

## Stand Initialization

```{r fvs-stand}
# Combine plot summary with site data
plots <- get_plot_summary(trees)
site <- get_site_data(fia_local)
stand_data <- merge(plots, site, by = "plot_cn")

export_fvs_standlist(stand_data, "stands.csv", variant = "NE")
```

# 8. Time Series & Remeasurement

## Find Remeasured Plots

```{r remeasured}
remeas <- get_remeasured_plots(fia)
head(remeas)
# plot_cn, n_measurements, first_year, last_year, years

# Plots measured 3+ times
remeas[remeas$n_measurements >= 3, ]
```

## Build Time Series

```{r timeseries}
ts <- build_time_series(fia, include_dead = TRUE)
head(ts)
# All build_tree_list() columns PLUS:
# measurement_number, years_since_previous,
# prev_dbh, prev_height, dbh_growth, height_growth,
# annual_dbh_growth, ingrowth, mortality
```

## Growth Summaries

```{r growth}
growth <- summarize_growth(ts)
head(growth)
# plot_key, period_start, period_end, period_years,
# net_ba_growth_per_acre, gross_ba_growth_per_acre,
# mortality_ba_per_acre, ingrowth_tpa,
# annual_dbh_growth, mortality_rate

# Average net BA growth
mean(growth$net_ba_growth_per_acre, na.rm = TRUE)
# e.g., 1.2 ft²/ac/yr
```

# 9. Growth Modeling

The time series and competition functions produce analysis-ready data
for individual tree growth models.

## Prepare Model Data

```{r model-prep}
# Build time series with competition indices
ts <- build_time_series(fia, include_dead = TRUE)
panel <- get_individual_tree_history(ts, min_measurements = 2)

# Key columns for growth modeling:
# dbh, annual_dbh_growth (response variable)
# spp_code, bal, plot_ba, relative_dbh (predictors)
# site_index, slope, aspect, elevation (site variables)
```

## Fit a Diameter Increment Model

```{r model-fit}
# Filter to live survivors with valid growth
model_data <- panel |>
  dplyr::filter(
    status == "live",
    !ingrowth,
    !is.na(annual_dbh_growth),
    annual_dbh_growth > 0,
    annual_dbh_growth < 1.0  # Remove obvious errors
  )

# Simple linear model
mod <- lm(
  log(annual_dbh_growth) ~ log(dbh) + bal + spp_code + site_index,
  data = model_data
)
summary(mod)
```

## Mortality Model

```{r mortality}
# Binary mortality: did the tree survive the period?
mort_data <- panel |>
  dplyr::filter(
    measurement_number > 1,
    !ingrowth
  ) |>
  dplyr::mutate(died = as.integer(mortality))

mort_model <- glm(
  died ~ dbh + bal + spp_code,
  family = binomial,
  data = mort_data
)
summary(mort_model)
```

## Enrich with Climate and Terrain

```{r enrich}
# Get spatial covariates for model data
plots_sf <- plot_locations(fia)
plots_sf <- get_terrain_data(plots_sf)
plots_sf <- get_climate_data(plots_sf, years = 2019)
plots_sf <- get_soils_data(plots_sf)

# Join back to model data
# (merge on plot coordinates)
```

# 10. Utility Functions

## Species Lookups

```{r species-util}
# FIA code → FVS code
fia_to_fvs(c(833, 12, 129))
# "RO" "BF" "WP"

# FVS code → FIA code
fvs_to_fia(c("RO", "BF", "WP"))
# 833  12 129

# Common names
get_common_name(c(833, 12, 129))
# "northern red oak" "balsam fir" "eastern white pine"

# Full species table
spp_table <- get_species_table("NE")
```

## Column Name Translation

```{r dictionary}
# If you have raw FIA data and want forestry names
df <- data.frame(DIA = 10, HT = 65, STATUSCD = 1)
fia_to_forestry_names(df)
# dbh = 10, height = 65, status_code = 1

# Reverse (for interfacing with other FIA tools)
forestry_to_fia_names(df_forestry)
```

## Outlier Detection

```{r outliers}
flagged <- flag_outliers(trees)
# Adds columns: warn_dbh, flag_dbh, outlier_reason, etc.

# See flagged trees
flagged[!is.na(flagged$outlier_reason), c("species", "dbh", "height", "outlier_reason")]
```

## Regeneration

```{r regen}
regen <- summarize_regen(
  stand = trees$plot_id,
  plot = trees$plot_cn,
  subplot = trees$subplot,
  dbh = trees$dbh,
  species = trees$spp_code
)
head(regen)
# regen_tpa, regen_ba_per_acre, stocking_pct
```

## Variance Estimation

```{r variance}
# Standard error for BA/acre estimate
plots <- get_plot_summary(trees)
est <- estimate_variance(
  stand = rep(1, nrow(plots)),
  plot = plots$plot_id,
  y = plots$ba_per_acre,
  conf_level = 0.90
)
est$mean       # Mean BA/acre
est$se         # Standard error
est$ci_lower   # 90% CI lower
est$ci_upper   # 90% CI upper
est$precision_pct  # Precision as % of mean
```

---

# Appendix A: End-to-End Workflow — Property Inventory

A forester receives a shapefile of a client's property and needs to produce
FVS-ready tree lists and stand summaries.

```{r workflow-a}
library(FIAwrapper)
set_fvs_variant("NE")

# 1. Get tree list from shapefile
trees <- get_tree_list(
  states = "ME",
  aoi = "client_property.shp",
  include_dead = TRUE
)

# 2. Review data quality
flagged <- flag_outliers(trees)
issues <- flagged[!is.na(flagged$outlier_reason), ]
message(nrow(issues), " trees flagged for review")

# 3. Plot summaries
plots <- get_plot_summary(trees)
print(plots[, c("plot_id", "tpa", "ba_per_acre", "qmd", "dominant_species")])

# 4. Species composition
spp <- get_species_summary(trees)

# 5. Site data
site <- get_site_data(fetch_fia_data("ME"))

# 6. Export to FVS
export_fvs_treelist(trees, "client_fvs.tre", variant = "NE")
export_fvs_standlist(merge(plots, site, by = "plot_cn"), "client_stands.csv")

message("Done! FVS files ready for simulation.")
```

# Appendix B: End-to-End Workflow — Regional Growth Analysis

A researcher wants to analyze diameter growth patterns across Maine using
FIA remeasurement data.

```{r workflow-b}
library(FIAwrapper)
set_fvs_variant("NE")

# 1. Download state data
fia <- fetch_fia_data("ME", data_dir = "~/fia_data")

# 2. Find remeasured plots
remeas <- get_remeasured_plots(fia)
message(nrow(remeas), " plots with multiple measurements")

# 3. Build time series
ts <- build_time_series(fia, include_dead = TRUE)

# 4. Get panel data with competition
panel <- get_individual_tree_history(ts, min_measurements = 2)

# 5. Filter to quality growth data
model_data <- panel |>
  dplyr::filter(
    status == "live",
    !ingrowth,
    !is.na(annual_dbh_growth),
    annual_dbh_growth > 0,
    annual_dbh_growth < 1.0
  )

# 6. Fit growth model
mod <- lm(
  log(annual_dbh_growth) ~ log(dbh) + I(dbh^2) + bal + spp_code,
  data = model_data
)
summary(mod)

# 7. Summarize growth by species
growth_by_spp <- model_data |>
  dplyr::group_by(spp_code) |>
  dplyr::summarize(
    n_trees = dplyr::n(),
    mean_growth = mean(annual_dbh_growth),
    sd_growth = sd(annual_dbh_growth),
    mean_dbh = mean(dbh)
  ) |>
  dplyr::arrange(dplyr::desc(n_trees))

print(growth_by_spp, n = 20)
```

# Appendix C: End-to-End Workflow — Species Composition Change

Tracking how species composition shifts over time in a given area.

```{r workflow-c}
library(FIAwrapper)
set_fvs_variant("NE")

# 1. Build time series
fia <- fetch_fia_data("VT", data_dir = "~/fia_data")
ts <- build_time_series(fia, include_dead = FALSE)

# 2. Summarize species BA by measurement period
spp_time <- ts |>
  dplyr::filter(status == "live") |>
  dplyr::group_by(inventory_year, spp_code) |>
  dplyr::summarize(
    mean_ba_per_acre = mean(ba_per_acre, na.rm = TRUE),
    mean_tpa = mean(tpa, na.rm = TRUE),
    .groups = "drop"
  )

# 3. Compare softwood vs hardwood trends
spp_table <- get_species_table("NE")
spp_time <- dplyr::left_join(spp_time, spp_table[, c("fvs_code", "wood_type")],
                               by = c("spp_code" = "fvs_code"))

wood_trend <- spp_time |>
  dplyr::group_by(inventory_year, wood_type) |>
  dplyr::summarize(total_ba = sum(mean_ba_per_acre, na.rm = TRUE),
                    .groups = "drop")

print(wood_trend)

# 4. Plot trend
if (requireNamespace("ggplot2", quietly = TRUE)) {
  ggplot2::ggplot(wood_trend,
    ggplot2::aes(x = inventory_year, y = total_ba, color = wood_type)) +
    ggplot2::geom_line(linewidth = 1.2) +
    ggplot2::labs(x = "Inventory Year", y = "BA per Acre (ft²/ac)",
                   title = "Vermont: Softwood vs Hardwood BA Over Time")
}
```

# Appendix D: FIA → Forestry Column Name Reference

| FIA Column | Forestry Name | Description |
|------------|--------------|-------------|
| DIA | dbh | Diameter at breast height (inches) |
| HT | height | Total height (feet) |
| CR | crown_ratio | Crown ratio (%) |
| STATUSCD | status_code | 1=live, 2=dead, 3=removed |
| SPCD | spcd | FIA numeric species code |
| INVYR | inventory_year | Year of measurement |
| STATECD | state_code | FIPS state code |
| COUNTYCD | county_code | FIPS county code |
| LAT / LON | lat / lon | Fuzzed coordinates |
| ELEV | elevation | Elevation (feet) |
| SLOPE | slope | Slope (%) |
| ASPECT | aspect | Aspect (degrees) |
| FORTYPCD | forest_type_code | FIA forest type |
| STDAGE | stand_age | Stand age (years) |
| SICOND | site_index | Condition-level site index |
| SITECLCD | site_class | Site productivity class |
| CONDPROP_UNADJ | cond_proportion | Condition proportion (0-1) |
| TPA_UNADJ | tpa_unadj | Unadjusted TPA from FIA |
| VOLCFNET | vol_cf_net | Net cubic foot volume |
| VOLBFNET | vol_bf_net | Net board foot volume |
| DRYBIO_AG | biomass_ag_dry | Aboveground dry biomass |
| CARBON_AG | carbon_ag | Aboveground carbon |
| SDI | stand_density_index | Stand density index |
| BALIVE | ba_live | Live tree BA on condition |
| CN | tree_cn | Unique tree control number |
| PREV_TRE_CN | prev_tree_cn | Previous measurement CN |
| PLT_CN | plot_cn | Plot control number |
| DECAYCD | decay_class | Decay class (dead trees) |
| CCLCD | crown_class | Crown class code |

Use `fia_to_forestry_names(df)` and `forestry_to_fia_names(df)` to convert
column names programmatically.
