---
title: "FIAwrapper: Complete Package Guide"
subtitle: "Clean FIA Tree Lists in Forestry Language"
author: "Ryan Smith"
date: last-modified
format:
  html:
    toc: true
    toc-depth: 4
    toc-expand: 2
    number-sections: true
    code-fold: false
    code-overflow: wrap
    theme: flatly
    highlight-style: github
    anchor-sections: true
    smooth-scroll: true
    link-external-newwindow: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  eval = FALSE,
  echo = TRUE,
  message = TRUE,
  warning = FALSE,
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction {.unnumbered}

**FIAwrapper** transforms raw USDA Forest Inventory and Analysis (FIA) database
tables into clean, imperial-unit tree lists using standard forestry terminology.
No more deciphering `CONDPROP_UNADJ`, `TPA_UNADJ`, or `SPCD` --- just `dbh`,
`height`, `tpa`, and `species`.

The package is designed for foresters, silviculturists, and forest biometricians
who need to:

- Build **per-acre tree lists** from FIA data for any area of interest
- Export tree lists directly to the **Forest Vegetation Simulator (FVS)**
- Analyze **remeasurement data** for growth modeling, mortality analysis, and
  stand dynamics
- Summarize **plot-level and species-level** attributes in familiar units
- Enrich plot data with **terrain, climate, and soils** from federal APIs
- Flag **outlier trees** using biologically plausible bounds
- Estimate **sampling error** for any attribute

Every function in FIAwrapper uses **imperial units** by default: DBH in inches,
height in feet, basal area in ft²/acre, and trees per acre (TPA). FIA database
column names like `DIA`, `HT`, `STATUSCD`, and `PLT_CN` are translated to
forestry terms like `dbh`, `height`, `status`, and `plot_cn`.

## How This Guide Is Organized

| Section | What You'll Learn |
|---------|-------------------|
| @sec-install | Installation, FVS variant setup, Python environment |
| @sec-fia-background | FIA program, plot design, table structure |
| @sec-data-acquisition | Downloading and loading FIA data |
| @sec-spatial | Filtering plots by shapefile or bounding box |
| @sec-tree-lists | Building clean tree lists from raw FIA data |
| @sec-plot-site | Plot summaries, site attributes, species composition |
| @sec-fvs | Exporting to FVS formats |
| @sec-time-series | Remeasurement, growth tracking, mortality |
| @sec-competition | Competition indices for growth modeling |
| @sec-quality | Outlier detection, variance estimation, regeneration |
| @sec-species | Species code system and lookups |
| @sec-utilities | Unit conversions, column name translation |
| @sec-spatial-apis | Terrain, climate, and soils from federal APIs |
| @sec-workflows | End-to-end workflows for common goals |

## Example Dataset

This guide uses `fia_ri`, a realistic FIA dataset bundled with the package.
It contains 12 inventory plots across 3 Rhode Island counties with 2
measurement cycles (2014 and 2019), representing typical southern New England
mixed forests. See @sec-example-data for details.

```{r}
library(FIAwrapper)
data(fia_ri)

names(fia_ri)
#> [1] "TREE" "PLOT" "COND"

nrow(fia_ri$TREE)
#> [1] 467

table(fia_ri$PLOT$INVYR)
#> 2014 2019
#>    6   12
```

# Installation & Setup {#sec-install}

## Installing the Package

```{r}
# Install from GitHub
devtools::install_github("ryanmismith/FIAwrapper")

library(FIAwrapper)
```

**Dependencies**: FIAwrapper imports `dplyr`, `tibble`, `rlang`, and `readr`.
Optional packages include `rFIA` (for data downloads), `sf` (for spatial
operations), `httr2` (for DataMart downloads), and `reticulate` (for Python
spatial backend).

## Setting Your FVS Variant {#sec-variant}

The FVS variant determines which species code mapping is used throughout the
session. Every function that works with species codes respects this setting.

```{r}
set_fvs_variant("NE")  # Northeast variant
#> FVS variant set to: NE
```

**Available variants** (see the [FVS website](https://www.fs.usda.gov/fvs/whatis.shtml)
for coverage maps):

| Code | Variant | Region |
|------|---------|--------|
| NE | Northeast | New England, Mid-Atlantic, Lake States |
| SN | Southern | Southeast US, Gulf states |
| LS | Lake States | WI, MI, MN |
| CS | Central States | Central hardwoods |
| PN | Pacific Northwest — Coast | Western OR, WA |
| WC | West Cascades | Cascade Range |
| BM | Blue Mountains | Eastern OR, WA |
| CR | Central Rockies | CO, eastern WY |
| EM | Eastern Montana | MT east of divide |
| IE | Inland Empire | Northern ID, western MT |
| SO | Southern Oregon | Southern OR |
| AK | Alaska | AK |

If you don't set a variant, FIAwrapper uses a default mapping table that covers
common species across regions. You can check the current variant at any time:

```{r}
getOption("FIAwrapper.variant")
#> [1] "NE"
```

::: {.callout-tip}
## When to Set a Variant
Set the variant **before** building tree lists or exporting to FVS. The variant
determines the 2-letter alpha code (e.g., "RM" for red maple in NE vs. "RM" in
SN) used in the `spp_code` column of your tree list.
:::

## Python Environment (Optional) {#sec-python-env}

Spatial functions (`get_plots_in_area()`, `get_terrain_data()`,
`get_climate_data()`, `get_soils_data()`) use Python libraries (geopandas,
shapely) behind the scenes via `reticulate`. The conda environment is created
automatically on first use.

```{r}
# Check if Python environment is ready
check_python_deps()
#> TRUE

# Manually create/update the environment
ensure_python_env()
```

If you don't need spatial functions, Python is never required. The core tree
list, summary, and FVS export functions are pure R.

::: {.callout-note}
## sf as Primary Spatial Engine
FIAwrapper uses the R `sf` package as the primary spatial engine when available.
Python (geopandas) is a fallback for environments where `sf` is hard to install.
Most users will never need the Python backend.
:::

# Understanding FIA Data {#sec-fia-background}

## The FIA Program

The USDA Forest Service [Forest Inventory and Analysis](https://www.fia.fs.usda.gov/)
(FIA) program is the nation's continuous forest census. FIA maintains a network
of ~320,000 permanent sample plots across all 50 states, remeasured on a
rotating 5-to-10-year cycle.

**Key references**:

- [FIADB User Guide](https://www.fia.fs.usda.gov/library/database-documentation/)
  --- Complete database documentation
- [FIA DataMart](https://apps.fs.usda.gov/fia/datamart/) --- Download raw data
- [EVALIDator](https://apps.fs.usda.gov/fiadb-api/evalidator) --- Web tool for
  pre-computed estimates
- Bechtold & Patterson (2005). *The Enhanced FIA Program — National Consistency
  with Regional Relevance*. Gen. Tech. Rep. SRS-80.
- Burrill et al. (2024). *The Forest Inventory and Analysis Database: Database
  Description and User Guide for Phase 2*. Version 9.2.

## FIA Plot Design {#sec-plot-design}

Each FIA plot is a **cluster of 4 subplots** arranged in a fixed pattern.
Understanding this design is critical for correct per-acre expansion.

```
     Subplot 2
        o
        |  120 ft
        |
Subplot 3 ---o--- Subplot 4
        |  120 ft (horizontal)
        |
     Subplot 1
     (plot center)
```

Each subplot contains two concentric sampling areas:

| Sample Area | Radius | Area | Trees Measured |
|------------|--------|------|----------------|
| **Subplot** | 24.0 ft (7.32 m) | 1,809.6 ft² | All trees ≥ 5.0 in DBH |
| **Microplot** | 6.8 ft (2.07 m) | 145.3 ft² | Saplings 1.0–4.9 in DBH |

**Per-acre expansion factors** (TPA) for a standard 4-subplot plot:

- Subplot trees (≥ 5.0 in DBH): `TPA = 1 / (π × 24² / 43560 × 4) ≈ 6.018`
- Microplot trees (< 5.0 in DBH): `TPA = 1 / (π × 6.8² / 43560 × 4) ≈ 74.965`

These expansion factors convert individual tree measurements to per-acre values.
FIAwrapper computes them automatically via `compute_tpa()` or within
`build_tree_list()`.

::: {.callout-important}
## Condition Proportion
When a plot straddles two forest conditions (e.g., half oak forest, half field),
FIA assigns a `CONDPROP_UNADJ` (0 to 1) to each condition. The TPA calculation
must account for this: `TPA_adjusted = TPA_base / CONDPROP_UNADJ`. FIAwrapper
handles this automatically when `CONDPROP_UNADJ` is available in the COND table.
:::

## FIA Database Tables {#sec-fia-tables}

FIA data is organized into relational tables. FIAwrapper uses three core tables:

### TREE Table
One row per tree per measurement. Key columns:

| FIA Column | FIAwrapper Name | Description |
|-----------|----------------|-------------|
| CN | tree_cn | Unique tree identifier |
| PLT_CN | plot_cn | Foreign key to PLOT table |
| SUBP | subplot | Subplot number (1–4) |
| STATUSCD | status_code | 1 = live, 2 = dead, 3 = removed |
| SPCD | spcd | FIA numeric species code (see [REF_SPECIES](https://apps.fs.usda.gov/fia/datamart/CSV/REF_SPECIES.csv)) |
| DIA | dbh | Diameter at breast height (inches) |
| HT | height | Total tree height (feet) |
| CR | crown_ratio | Compacted crown ratio (percent) |
| TPA_UNADJ | tpa_unadj | Unadjusted trees per acre |
| PREV_TRE_CN | prev_tree_cn | CN of same tree in previous measurement |
| VOLCFNET | vol_cf_net | Net cubic foot volume |
| VOLBFNET | vol_bf_net | Net board foot volume (Scribner) |
| DRYBIO_AG | biomass_ag_dry | Aboveground dry biomass (lbs) |
| CARBON_AG | carbon_ag | Aboveground carbon (lbs) |

### PLOT Table
One row per plot per measurement. Key columns:

| FIA Column | FIAwrapper Name | Description |
|-----------|----------------|-------------|
| CN | plot_cn | Unique plot identifier (use as join key) |
| STATECD | state_code | FIPS state code |
| COUNTYCD | county_code | FIPS county code |
| PLOT | plot_number | Plot number within county |
| INVYR | inventory_year | Year of measurement |
| LAT / LON | lat / lon | Approximate coordinates (fuzzed ±0.5–1 mi) |
| ELEV | elevation | Elevation (feet) |
| PREV_PLT_CN | — | CN of same plot in previous measurement |

::: {.callout-warning}
## Fuzzed Coordinates
FIA intentionally offsets plot coordinates by up to 1 mile for privacy. These
coordinates are adequate for regional analysis and spatial filtering, but should
**not** be used for precise site-level work. Exact coordinates require a
[confidential data agreement](https://www.fia.fs.usda.gov/tools-data/spatial/index.php).
:::

### COND Table
One row per condition per plot per measurement. Key columns:

| FIA Column | FIAwrapper Name | Description |
|-----------|----------------|-------------|
| PLT_CN | plot_cn | Foreign key to PLOT |
| CONDID | condition_id | Condition identifier within plot |
| CONDPROP_UNADJ | cond_proportion | Proportion of plot in this condition (0–1) |
| FORTYPCD | forest_type_code | Forest type code (see [REF_FOREST_TYPE](https://apps.fs.usda.gov/fia/datamart/CSV/REF_FOREST_TYPE.csv)) |
| STDAGE | stand_age | Stand age (years) |
| SICOND | site_index | Site index (base age 50, feet) |
| SITECLCD | site_class | Site productivity class (1–7) |
| SLOPE | slope | Slope in percent |
| ASPECT | aspect | Aspect in degrees (0–360) |
| BALIVE | ba_live | Live tree basal area (ft²/ac) on condition |
| OWNGRPCD | ownership_group | Ownership group code |

## How FIAwrapper Connects the Tables {#sec-table-joins}

When you call `build_tree_list()`, FIAwrapper performs these joins automatically:

1. **TREE → PLOT**: via `PLT_CN` to get location, year, and state info
2. **TREE → COND**: via `PLT_CN` + `CONDID` to get site attributes
3. **Species translation**: `SPCD` → `spp_code` (FVS alpha) and `species`
   (common name) using the active variant mapping
4. **TPA computation**: using `compute_tpa()` with the FIA subplot/microplot
   design
5. **Column renaming**: FIA jargon → forestry terms via `fia_to_forestry_names()`

The result is a single tibble with one row per tree and all measurements in
imperial units.

## The Example Dataset {#sec-example-data}

FIAwrapper ships with `fia_ri`, a realistic dataset modeled on Rhode Island
FIA data. It follows FIADB v9.2 specifications exactly and contains:

- **467 tree records** across 12 plots and 3 counties
- **2 inventory cycles** (2014, 2019) with 6 remeasured plots
- **15 species** typical of southern New England mixed forest
- Standing dead trees (~5% of stems) for mortality analysis
- Saplings on microplots for regeneration assessment

```{r}
data(fia_ri)

# Structure
str(fia_ri, max.level = 1)
#> List of 3
#>  $ TREE:'data.frame': 467 obs. of 22 variables
#>  $ PLOT:'data.frame': 18 obs. of 11 variables
#>  $ COND:'data.frame': 18 obs. of 16 variables

# Species in the dataset
sort(unique(fia_ri$TREE$SPCD))
#>  [1]  12  68  71  97 129 241 261 316 318 531 541 621 802 833 951

# What are these species?
set_fvs_variant("NE")
get_common_name(sort(unique(fia_ri$TREE$SPCD)))
```

This dataset is used in every example throughout this guide.

# Data Acquisition {#sec-data-acquisition}

FIAwrapper provides two functions for getting FIA data onto your machine:
`fetch_fia_data()` for downloading, and `load_fia_data()` for loading
previously downloaded data.

## Downloading FIA Data

`fetch_fia_data()` downloads FIA tables for one or more states. It tries the
[rFIA](https://cran.r-project.org/package=rFIA) package first (if installed),
then falls back to direct CSV downloads from the
[FIA DataMart](https://apps.fs.usda.gov/fia/datamart/).

```{r}
# Download core tables for Maine and New Hampshire
fia <- fetch_fia_data(
  states = c("ME", "NH"),
  tables = c("TREE", "PLOT", "COND", "SUBPLOT"),
  data_dir = "~/fia_data",
  method = "auto"  # tries rFIA first, then DataMart
)

names(fia)
#> [1] "TREE" "PLOT" "COND" "SUBPLOT"
```

**Parameters**:

- `states`: Character vector of 2-letter state abbreviations (e.g., `c("ME", "NH", "VT")`)
- `tables`: Which FIA tables to download. Default: `c("TREE", "PLOT", "COND", "SUBPLOT")`
- `data_dir`: Directory to cache downloads. Default: `getOption("FIAwrapper.data_dir")` or `tempdir()`
- `method`: `"auto"` (default), `"rfia"`, or `"datamart"`

**Caching**: Downloaded files are cached locally. Files less than 30 days old are
reused automatically. Delete the cache directory to force a fresh download.

::: {.callout-tip}
## rFIA vs DataMart
The `rFIA` method is faster and more reliable for multi-state downloads. Install
it with `install.packages("rFIA")`. The DataMart fallback downloads individual
CSV files from [apps.fs.usda.gov/fia/datamart/CSV/](https://apps.fs.usda.gov/fia/datamart/CSV/).
:::

## Loading Cached Data

```{r}
# Load previously downloaded data
fia <- load_fia_data("~/fia_data", states = "ME")
```

`load_fia_data()` auto-detects whether data was saved in rFIA format (RDS files)
or as raw CSVs. Optionally filter by state when loading multi-state datasets.

## Using the Example Dataset

For this guide, we use the bundled `fia_ri` dataset:

```{r}
data(fia_ri)
# This is equivalent to what fetch_fia_data() returns:
# a named list with TREE, PLOT, and COND data frames
```

# Spatial Queries {#sec-spatial}

FIAwrapper can filter FIA plots to a specific area of interest (AOI) using
shapefiles, `sf` objects, or bounding boxes.

## Filtering Plots by Area

`get_plots_in_area()` is the core spatial function. It finds all FIA plots
within your AOI and returns a filtered copy of the FIA data.

```{r}
# From a shapefile
fia_local <- get_plots_in_area(
  aoi = "my_property.shp",
  fia_data = fia,
  buffer_ft = 5280  # 1-mile buffer around the AOI
)

# From an sf object
library(sf)
my_bbox <- st_as_sf(st_as_sfc(st_bbox(
  c(xmin = -71.6, ymin = 41.4, xmax = -71.3, ymax = 41.9),
  crs = 4326
)))
fia_local <- get_plots_in_area(aoi = my_bbox, fia_data = fia_ri)

# From a bounding box vector (xmin, ymin, xmax, ymax)
fia_local <- get_plots_in_area(
  aoi = c(-71.6, 41.4, -71.3, 41.9),
  fia_data = fia_ri
)
```

**Parameters**:

- `aoi`: File path (shapefile, GeoPackage, GeoJSON), `sf` object, or numeric
  vector `c(xmin, ymin, xmax, ymax)` in WGS84
- `fia_data`: FIA data list (from `fetch_fia_data()`) or character vector of
  state abbreviations (auto-downloads)
- `buffer_ft`: Buffer distance in feet around the AOI (default: 0)
- `data_dir`: Cache directory when `fia_data` is state abbreviations

**Returns**: A filtered FIA data list (same structure as input) containing only
plots within the AOI. All tables (TREE, PLOT, COND) are filtered consistently.

::: {.callout-note}
## Spatial Engine
`get_plots_in_area()` uses the R `sf` package when available. If `sf` is not
installed, it falls back to Python geopandas via `reticulate`. Install `sf`
with `install.packages("sf")` for the simplest setup.
:::

## Extracting Plot Coordinates

`plot_locations()` creates an `sf` POINT dataset from FIA plot coordinates.

```{r}
plots_sf <- plot_locations(fia_ri)
class(plots_sf)
#> [1] "sf"         "data.frame"

head(plots_sf[, c("plot_cn", "inventory_year")])
```

**Returns**: An `sf` data frame with POINT geometry in the specified CRS
(default: EPSG:4326 / WGS84). Includes all PLOT table columns.

## Mapping Plots

`map_plots()` creates a quick interactive or static map.

```{r}
# Basic map
map_plots(fia_ri)

# With AOI overlay
map_plots(fia_ri, aoi = "my_property.shp")
```

Uses `leaflet` for interactive maps (if installed), otherwise falls back to
base R `plot()`. The AOI boundary is overlaid when provided.

# Building Tree Lists {#sec-tree-lists}

## `build_tree_list()` — The Core Function {#sec-build-tree-list}

This is the main workhorse function. It transforms raw FIA tables into a clean,
analysis-ready tree list.

```{r}
set_fvs_variant("NE")

trees <- build_tree_list(
  fia_data = fia_ri,
  include_dead = FALSE,
  variant = "NE",
  most_recent = TRUE,
  compute_expansion = TRUE
)

nrow(trees)
#> [1] 452

head(trees[, c("plot_id", "spp_code", "species", "dbh", "height", "tpa", "ba_per_acre")])
```

**Parameters**:

- `fia_data`: Named list with TREE, PLOT, and COND data frames
- `include_dead`: Include standing dead trees? Default `FALSE`. When `TRUE`,
  adds a `status` column ("live" / "dead")
- `variant`: FVS variant code for species translation. Default: session variant
- `most_recent`: Keep only the latest inventory cycle? Default `TRUE`
- `compute_expansion`: Compute TPA from subplot geometry? Default `TRUE`.
  When `FALSE`, uses FIA's `TPA_UNADJ` directly

### Output Columns {#sec-tree-list-columns}

The output tibble contains one row per tree with these columns:

**Tree identity and location**:

| Column | Type | Description |
|--------|------|-------------|
| `plot_id` | character | Unique plot ID: `STATECD_COUNTYCD_PLOT` |
| `plot_cn` | integer | FIA plot control number |
| `subplot` | integer | Subplot number (1–4) |
| `tree_num` | integer | Tree number within subplot |
| `tree_cn` | integer | Unique tree control number |
| `inventory_year` | integer | Year of measurement |

**Species**:

| Column | Type | Description |
|--------|------|-------------|
| `species` | character | Common name (e.g., "red maple") |
| `spp_code` | character | FVS alpha code (e.g., "RM") |
| `spcd` | integer | FIA numeric code (e.g., 316), kept for reference |

**Measurements** (imperial units):

| Column | Type | Description |
|--------|------|-------------|
| `dbh` | numeric | Diameter at breast height (inches) |
| `height` | numeric | Total height (feet) |
| `crown_ratio` | numeric | Crown ratio (percent, 0–100) |
| `crown_class` | integer | Crown class (1=open-grown, 2=dominant, 3=codominant, 4=intermediate, 5=overtopped) |

**Per-acre values**:

| Column | Type | Description |
|--------|------|-------------|
| `tpa` | numeric | Trees per acre this tree represents |
| `ba_tree` | numeric | Individual tree basal area (ft²) |
| `ba_per_acre` | numeric | BA contribution = `ba_tree × tpa` (ft²/acre) |

**Status**:

| Column | Type | Description |
|--------|------|-------------|
| `status` | character | "live" or "dead" |
| `status_code` | integer | FIA STATUSCD (1, 2, or 3) |
| `decay_class` | integer | Decay class for dead trees (1–5) |
| `damage_code1` | integer | Primary damage agent code |

**Volume and biomass**:

| Column | Type | Description |
|--------|------|-------------|
| `vol_cf_net` | numeric | Net cubic foot volume |
| `vol_cf_gross` | numeric | Gross cubic foot volume |
| `vol_bf_net` | numeric | Net board foot volume (Scribner) |
| `biomass_ag_dry` | numeric | Aboveground dry biomass (lbs) |
| `carbon_ag` | numeric | Aboveground carbon (lbs) |

**Site and location**:

| Column | Type | Description |
|--------|------|-------------|
| `state_code` | integer | FIPS state code |
| `county_code` | integer | FIPS county code |
| `lat` / `lon` | numeric | Fuzzed coordinates |
| `elevation` | numeric | Elevation (feet) |
| `slope` | numeric | Slope (percent) |
| `aspect` | numeric | Aspect (degrees) |
| `site_index` | numeric | Site index (base age 50, feet) |
| `site_class` | integer | Site productivity class |
| `stand_age` | integer | Stand age (years) |
| `forest_type_code` | integer | FIA forest type code |

### How TPA Is Computed {#sec-tpa-computation}

`build_tree_list()` computes TPA using the FIA nested plot design:

1. Trees with **DBH ≥ 5.0 inches** are measured on the **subplot** (24 ft
   radius). Each subplot covers `π × 24² / 43560 = 0.04154 acres`. With 4
   subplots, each tree represents `1 / (0.04154 × 4) ≈ 6.018 trees/acre`.

2. Trees with **DBH < 5.0 inches** are measured on the **microplot** (6.8 ft
   radius). Each microplot covers `π × 6.8² / 43560 = 0.003336 acres`. With
   4 microplots, each tree represents `1 / (0.003336 × 4) ≈ 74.965 trees/acre`.

3. If `CONDPROP_UNADJ` < 1.0 (plot straddles conditions), TPA is adjusted:
   `TPA = TPA_base / CONDPROP_UNADJ`.

::: {.callout-important}
## TPA vs TPA_UNADJ
FIA provides `TPA_UNADJ` in the TREE table, which is their own expansion
factor. FIAwrapper's `compute_tpa()` recalculates this from first principles
using the subplot geometry. In most cases the values agree closely. Set
`compute_expansion = FALSE` to use FIA's values instead.
:::

### Working with Dead Trees

```{r}
all_trees <- build_tree_list(fia_ri, include_dead = TRUE)

# Status breakdown
table(all_trees$status)
#>  dead  live
#>    15   452

# Dead tree attributes
snags <- all_trees[all_trees$status == "dead", ]
head(snags[, c("spp_code", "dbh", "height", "decay_class", "tpa")])
```

Dead trees (STATUSCD = 2) are standing dead / snags. They have valid DBH and
height measurements, plus a `decay_class` (1 = recently dead, 5 = heavily
decomposed). Mortality analysis requires `include_dead = TRUE`.

## `get_tree_list()` — One-Line Convenience

`get_tree_list()` chains `fetch_fia_data()` → `get_plots_in_area()` →
`build_tree_list()` into a single call.

```{r}
# Download + clip + build in one step
trees <- get_tree_list(
  states = "ME",
  aoi = "my_property.shp",
  data_dir = "~/fia_data",
  include_dead = TRUE,
  variant = "NE",
  most_recent = TRUE
)
```

**Parameters**: Combines all parameters from the three underlying functions.
Provide `states` and/or `aoi` to control data scope.

## Low-Level Data Prep: `prep_fia_data()`

For more control over the table joining process, use `prep_fia_data()` directly.
This function joins TREE, PLOT, and COND tables and renames columns, but does
**not** compute TPA or filter by status.

```{r}
prepped <- prep_fia_data(
  fia_trees = fia_ri$TREE,
  fia_plots = fia_ri$PLOT,
  fia_cond = fia_ri$COND,
  include_dead = TRUE,
  variant = "NE"
)

# All FIA columns renamed to forestry terms
names(prepped)
```

Useful when you want to apply custom filtering or compute expansion factors
differently (e.g., for non-standard plot designs).

# Plot & Site Data {#sec-plot-site}

## Plot Summaries {#sec-plot-summary}

`get_plot_summary()` aggregates a tree list to one row per plot with standard
forestry metrics.

```{r}
trees <- build_tree_list(fia_ri, variant = "NE")
plots <- get_plot_summary(trees)

head(plots[, c("plot_id", "tpa", "ba_per_acre", "qmd", "top_height",
               "species_count", "dominant_species")])
```

**Output columns**:

| Column | Units | Description |
|--------|-------|-------------|
| `tpa` | trees/acre | Total trees per acre |
| `ba_per_acre` | ft²/acre | Total basal area per acre |
| `qmd` | inches | Quadratic mean diameter |
| `mean_dbh` | inches | TPA-weighted mean DBH |
| `top_height` | feet | Mean height of tallest 40 TPA trees (site height proxy) |
| `mean_height` | feet | TPA-weighted mean height |
| `species_count` | count | Number of species on plot |
| `dominant_species` | character | Species with highest BA |
| `vol_per_acre` | ft³/acre | Net cubic foot volume (if available) |
| `n_trees` | count | Number of measured trees |

**Dead tree handling**: By default, `get_plot_summary()` excludes dead trees.
Set `include_dead = TRUE` to include them in all calculations.

```{r}
# Compare live-only vs all-stems summaries
plots_live <- get_plot_summary(all_trees, include_dead = FALSE)
plots_all  <- get_plot_summary(all_trees, include_dead = TRUE)

# Dead trees add to TPA and BA
mean(plots_all$tpa - plots_live$tpa)
```

::: {.callout-tip}
## QMD Computation
QMD (quadratic mean diameter) is computed as:
$$QMD = \sqrt{\frac{\sum_{i=1}^{n} DBH_i^2 \times TPA_i}{\sum_{i=1}^{n} TPA_i}}$$
This is the diameter of the tree with average basal area, and is always
larger than the arithmetic mean diameter. It's the standard metric used in
silvicultural prescriptions and yield tables.
:::

## Site Attributes {#sec-site-data}

`get_site_data()` extracts plot-level site and stand characteristics from the
PLOT and COND tables.

```{r}
site <- get_site_data(fia_ri)

head(site[, c("plot_cn", "inventory_year", "elevation", "slope", "aspect",
              "stand_age", "site_index", "forest_type_code")])
```

**Output columns**:

| Column | Units | Description |
|--------|-------|-------------|
| `elevation` | feet | Elevation above sea level |
| `slope` | percent | Slope gradient |
| `aspect` | degrees | Aspect (0 = north, 90 = east, 180 = south, 270 = west) |
| `forest_type_code` | integer | FIA forest type (see [REF_FOREST_TYPE](https://apps.fs.usda.gov/fia/datamart/CSV/REF_FOREST_TYPE.csv)) |
| `stand_age` | years | Stand age |
| `site_index` | feet | Site index at base age 50 (height of dominant trees) |
| `site_class` | integer | Site productivity class (1=highest, 7=lowest) |
| `ownership_group` | integer | Ownership code ([ref](https://www.fia.fs.usda.gov/library/database-documentation/)) |
| `stand_density_index` | index | Reineke's stand density index |
| `ba_live` | ft²/acre | Live basal area on condition |

**Merging with plot summaries**:

```{r}
# Combine metrics with site data
full_plots <- merge(plots, site, by = "plot_cn", suffixes = c("", "_site"))

# Now you have BA/acre + site index + slope + etc. in one table
head(full_plots[, c("plot_id", "ba_per_acre", "qmd", "site_index", "slope", "stand_age")])
```

## Species Composition {#sec-species-summary}

`get_species_summary()` creates per-species summaries for each plot.

```{r}
spp <- get_species_summary(trees)

head(spp[, c("plot_id", "spp_code", "species_name", "tpa", "ba_per_acre", "qmd", "pct_ba")])
```

**Output columns**:

| Column | Description |
|--------|-------------|
| `spp_code` | FVS alpha species code |
| `species_name` | Common name |
| `tpa` | Trees per acre for this species |
| `ba_per_acre` | Basal area (ft²/acre) for this species |
| `qmd` | QMD for this species (inches) |
| `mean_dbh` | Weighted mean DBH (inches) |
| `pct_ba` | Percent of total plot BA from this species |
| `n_trees` | Number of measured trees |

**Finding dominant species across all plots**:

```{r}
# Top species by mean BA/acre across all plots
spp |>
  dplyr::group_by(spp_code, species_name) |>
  dplyr::summarize(
    mean_ba = mean(ba_per_acre, na.rm = TRUE),
    plots_present = dplyr::n(),
    .groups = "drop"
  ) |>
  dplyr::arrange(dplyr::desc(mean_ba)) |>
  head(10)
```

### Relationship Between Summary Functions

```
build_tree_list()         # Tree-level data (1 row per tree)
    │
    ├── get_plot_summary()    # Plot-level (1 row per plot)
    │       Aggregates: TPA, BA/ac, QMD, species count
    │
    ├── get_species_summary() # Species×Plot level
    │       Aggregates: TPA, BA/ac, QMD per species per plot
    │
    └── get_site_data()       # Plot-level site attributes
            Source: PLOT + COND tables directly
```

## `summarize_plot()` — Vector-Based Alternative

For users who prefer working with raw vectors (not tibbles), `summarize_plot()`
accepts stand, plot, DBH, height, and TPA as separate vectors.

```{r}
result <- summarize_plot(
  stand   = trees$plot_id,
  plot    = trees$plot_cn,
  dbh     = trees$dbh,
  height  = trees$height,
  tpa     = trees$tpa,
  species = trees$spp_code
)
head(result)
```

This function also supports custom aggregation columns via `sum_cols` and
`mean_cols` parameters.

# FVS Export {#sec-fvs}

The [Forest Vegetation Simulator](https://www.fs.usda.gov/fvs/whatis.shtml)
(FVS) is the USDA Forest Service's standard growth and yield projection system.
FIAwrapper exports tree lists and stand initialization data in formats that FVS
can read directly.

**Key FVS references**:

- [FVS Overview](https://www.fs.usda.gov/fvs/whatis.shtml) — What FVS does
- [FVS Variants](https://www.fs.usda.gov/fvs/software/plantnat.php) — Regional
  model variants and species coverage
- [Suppose User Guide](https://www.fs.usda.gov/fvs/documents/userguides.shtml)
  — GUI for running FVS
- Dixon (2002). *Essential FVS: A User's Guide to the Forest Vegetation
  Simulator*. USDA-FS Internal Report.

## Setting the FVS Variant

```{r}
set_fvs_variant("NE")
#> FVS variant set to: NE

# The variant stays active for the session
getOption("FIAwrapper.variant")
#> [1] "NE"
```

The variant determines:

- Which 2-letter species codes are used (e.g., SPCD 316 = "RM" in NE, "RM" in SN)
- Which species mapping table is loaded from `inst/extdata/fvs_species_variants/`
- The `variant` field in stand initialization records

## Exporting Tree Lists {#sec-fvs-treelist}

`export_fvs_treelist()` writes a tree list in one of three FVS-compatible
formats.

```{r}
trees <- build_tree_list(fia_ri, variant = "NE")

# Fixed-width FVS format (.tre) — standard FVS input
export_fvs_treelist(trees, "fvs_input.tre", variant = "NE", format = "standard")

# CSV format — for Suppose or custom import
export_fvs_treelist(trees, "fvs_input.csv", variant = "NE", format = "csv")

# Suppose-compatible format
export_fvs_treelist(trees, "fvs_suppose.csv", variant = "NE", format = "suppose")
```

**Parameters**:

- `tree_list`: Tibble from `build_tree_list()` or similar
- `file`: Output file path
- `variant`: FVS variant code. Uses session variant if `NULL`
- `format`: `"standard"` (fixed-width .tre), `"csv"`, or `"suppose"`
- `stand_id_col`: Column for FVS stand ID (default: `"plot_id"`)

### FVS Standard Fixed-Width Format

The `"standard"` format writes the fixed-width tree record format documented in
the [FVS User Guide](https://www.fs.usda.gov/fvs/documents/userguides.shtml):

| Columns | Field | Source |
|---------|-------|--------|
| 1–4 | Stand ID | `plot_id` (truncated to 4 chars) |
| 5–7 | Tree ID | `tree_num` |
| 8–10 | Tree count (TPA) | `tpa` |
| 11–13 | History code | 1 = live, 6 = recent dead |
| 14–16 | Species code | FVS alpha code |
| 17–20 | DBH | Inches, 0.1 precision |
| 24–26 | Height | Feet, integer |
| 30–32 | Crown ratio code | Tenths (1–9) |

### CSV Format

The CSV format includes named columns: `stand_id`, `tree_id`, `species`, `dbh`,
`height`, `crown_ratio`, `tpa`, `status`.

### Suppose Format

The `"suppose"` format matches the column structure expected by the
[Suppose](https://www.fs.usda.gov/fvs/software/suppose.php) GUI: `StandID`,
`TreeID`, `TreeCount`, `History`, `Species`, `DBH`, `DG`, `Ht`, `HtTopK`,
`CrRatio`, damage codes, `TreeValue`, and `Prescription`.

## Exporting Stand Initialization {#sec-fvs-standlist}

`export_fvs_standlist()` writes stand-level data for FVS initialization.

```{r}
plots <- get_plot_summary(trees)
site  <- get_site_data(fia_ri)
stand_data <- merge(plots, site, by = "plot_cn")

export_fvs_standlist(stand_data, "stands.csv", variant = "NE")
```

**Output columns**: `stand_id`, `variant`, `latitude`, `longitude`, `elevation`,
`slope`, `aspect`, `site_index`, `ba_per_acre`, `tpa`.

::: {.callout-tip}
## FVS Workflow
A typical FVS workflow:

1. `build_tree_list()` → clean tree data
2. `get_plot_summary()` + `get_site_data()` → stand attributes
3. `export_fvs_treelist()` → `.tre` or `.csv` file
4. `export_fvs_standlist()` → stand initialization file
5. Load both files into FVS/Suppose and run projections
:::

# Time Series & Remeasurement {#sec-time-series}

FIA plots are remeasured on rotating cycles (typically 5–10 years). FIAwrapper
provides functions to identify remeasured plots, link trees across measurements,
and compute growth increments.

**Key concepts**:

- **PREV_PLT_CN**: Links the current plot measurement to the previous one
- **PREV_TRE_CN**: Links an individual tree record to its previous measurement
- **Ingrowth**: Trees that cross the 5.0-inch DBH threshold between measurements
- **Mortality**: Trees that transition from live (STATUSCD = 1) to dead (STATUSCD = 2)
- **Survivors**: Trees that remain live across both measurements

**References**:

- Shaw (2008). *Evaluation and Use of FIA Data for Growth and Yield
  Applications*. GTR-NRS-P-56.
- Westfall et al. (2023). *A National Assessment of Growth and Mortality from
  FIA Data*. Forest Science 69(3).

## Finding Remeasured Plots

`get_remeasured_plots()` identifies plots measured in multiple inventory cycles.

```{r}
remeas <- get_remeasured_plots(fia_ri)
remeas
#>   plot_cn state_code county_code n_measurements first_year last_year years
#> 1   10001         44           7              2       2014      2019 2014, 2019
#> 2   10002         44           7              2       2014      2019 2014, 2019
#> ...
```

**Returns**: Tibble with one row per unique plot location (identified by
`STATECD_COUNTYCD_PLOT`), showing measurement count and year range.

The function links plots across cycles using `PREV_PLT_CN` when available, or
falls back to matching on `STATECD` + `COUNTYCD` + `PLOT`.

## Building Time Series {#sec-build-time-series}

`build_time_series()` creates tree lists for ALL measurement cycles and links
individual trees across measurements.

```{r}
ts <- build_time_series(
  fia_ri,
  include_dead = TRUE,
  variant = "NE",
  min_measurements = 2
)

nrow(ts)
#> [1] 320

names(ts)
```

**Additional columns** beyond `build_tree_list()`:

| Column | Type | Description |
|--------|------|-------------|
| `plot_key` | character | Stable plot ID across measurements |
| `measurement_number` | integer | 1 = earliest, 2 = next, etc. |
| `years_since_previous` | numeric | Years between this and prior measurement |
| `prev_dbh` | numeric | DBH at previous measurement (inches) |
| `prev_height` | numeric | Height at previous measurement (feet) |
| `dbh_growth` | numeric | Periodic DBH increment (inches) |
| `height_growth` | numeric | Periodic height increment (feet) |
| `annual_dbh_growth` | numeric | Annualized DBH increment (inches/year) |
| `annual_height_growth` | numeric | Annualized height increment (feet/year) |
| `ingrowth` | logical | `TRUE` if tree was not present in prior measurement |
| `mortality` | logical | `TRUE` if tree died between measurements |

### How Trees Are Linked

1. **Primary**: Match via `PREV_TRE_CN` → `CN` (FIA's official tree link)
2. **Fallback**: Match on `plot_key` + `subplot` + `tree_num` when
   `PREV_TRE_CN` is not available

### Interpreting Growth Data

```{r}
# Survivor trees with valid growth
survivors <- ts |>
  dplyr::filter(
    measurement_number > 1,
    !ingrowth,
    !mortality,
    !is.na(dbh_growth)
  )

# Mean annual DBH increment
mean(survivors$annual_dbh_growth, na.rm = TRUE)

# Distribution by species
survivors |>
  dplyr::group_by(spp_code) |>
  dplyr::summarize(
    n = dplyr::n(),
    mean_ann_growth = mean(annual_dbh_growth, na.rm = TRUE),
    median_ann_growth = median(annual_dbh_growth, na.rm = TRUE)
  ) |>
  dplyr::arrange(dplyr::desc(n))
```

## Growth Summaries {#sec-growth-summary}

`summarize_growth()` aggregates time series data to plot-level growth metrics.

```{r}
growth <- summarize_growth(ts)
growth
```

**Output columns**:

| Column | Units | Description |
|--------|-------|-------------|
| `plot_key` | — | Stable plot identifier |
| `period_start` | year | Start of growth period |
| `period_end` | year | End of growth period |
| `period_years` | years | Length of period |
| `ba_per_acre_start` | ft²/ac | BA at start |
| `ba_per_acre_end` | ft²/ac | BA at end |
| `net_ba_growth_per_acre` | ft²/ac/yr | Net BA change per year |
| `gross_ba_growth_per_acre` | ft²/ac/yr | Gross BA growth (survivors + ingrowth) |
| `mortality_ba_per_acre` | ft²/ac/yr | BA lost to mortality per year |
| `ingrowth_tpa` | trees/ac/yr | New trees crossing threshold per year |
| `mean_dbh_growth` | in/yr | Mean annual DBH increment |
| `mortality_rate` | proportion | Fraction of stems that died |

**Net vs. Gross growth**:

- **Gross growth** = BA increment of survivors + BA of ingrowth trees
- **Mortality** = BA of trees that died during the period
- **Net growth** = Gross growth − Mortality

## Individual Tree History (Panel Data)

`get_individual_tree_history()` reshapes the time series into a panel format
ideal for longitudinal growth models.

```{r}
panel <- get_individual_tree_history(ts, min_measurements = 2)

# Each tree appears once per measurement with current + lagged values
head(panel[, c("tree_cn", "measurement_number", "dbh", "prev_dbh",
               "annual_dbh_growth", "spp_code")])
```

This function filters to trees measured at least `min_measurements` times and
adds competition metrics (BAL, plot BA, relative DBH) via `compute_competition()`.

# Competition Indices {#sec-competition}

Competition indices quantify the competitive environment experienced by
individual trees. They are essential predictors in diameter increment models,
mortality models, and crown ratio models.

**References**:

- Wykoff (1990). *A Basal Area Increment Model for Individual Conifers in the
  Northern Rocky Mountains*. Forest Science 36(4):1077-1104.
- Monserud & Sterba (1999). *Modeling Individual Tree Mortality for Austrian
  Forest Species*. Forest Ecology and Management 113:109-123.

## `compute_competition()`

Adds tree-level competition indices to any tree list.

```{r}
trees <- build_tree_list(fia_ri, variant = "NE")
comp <- compute_competition(trees)

# New columns added
setdiff(names(comp), names(trees))
#> [1] "plot_ba"  "plot_tpa"  "plot_qmd"  "bal"  "relative_dbh"  "percentile_dbh"
```

**Competition columns**:

| Column | Units | Description |
|--------|-------|-------------|
| `bal` | ft²/acre | **Basal area in larger trees** — sum of BA/acre from trees with DBH > this tree's DBH on the same plot. Key predictor of diameter growth. |
| `plot_ba` | ft²/acre | Total live basal area on the plot |
| `plot_tpa` | trees/acre | Total trees per acre on the plot |
| `plot_qmd` | inches | Plot-level quadratic mean diameter |
| `relative_dbh` | ratio | This tree's DBH ÷ plot QMD. Values > 1 indicate dominant trees. |
| `percentile_dbh` | 0–1 | Percentile rank of this tree's DBH within the plot |

### BAL (Basal Area in Larger Trees)

BAL is the most widely used individual-tree competition index. For a given tree,
it is the sum of per-acre basal area contributions from all trees on the same
plot that have a larger DBH.

$$BAL_i = \sum_{j: DBH_j > DBH_i} BA_j \times TPA_j$$

- Small, suppressed trees have high BAL (lots of larger trees above them)
- The largest tree on the plot has BAL = 0
- BAL is typically a strong negative predictor of diameter growth

```{r}
# BAL distribution
summary(comp$bal)

# Largest tree on each plot should have BAL = 0
comp |>
  dplyr::group_by(plot_id) |>
  dplyr::filter(dbh == max(dbh)) |>
  dplyr::pull(bal) |>
  unique()
#> [1] 0
```

### Using Competition in Growth Models

```{r}
# Prepare model data from time series
ts <- build_time_series(fia_ri, include_dead = TRUE, variant = "NE")
panel <- get_individual_tree_history(ts, min_measurements = 2)

# Filter to quality growth records
model_data <- panel |>
  dplyr::filter(
    status == "live",
    !ingrowth,
    !is.na(annual_dbh_growth),
    annual_dbh_growth > 0,
    annual_dbh_growth < 1.0  # cap at 1 in/yr
  )

# Fit diameter increment model
# Common form: ln(DG) = f(ln(DBH), BAL, species, site)
mod <- lm(
  log(annual_dbh_growth) ~ log(dbh) + bal + spp_code,
  data = model_data
)
summary(mod)
```

::: {.callout-tip}
## Classic Growth Model Form
The Wykoff (1990) diameter increment model uses:
$$\ln(DDS) = b_0 + b_1 \cdot DBH + b_2 \cdot DBH^2 + b_3 \cdot BAL + b_4 \cdot SI + b_5 \cdot CR + \epsilon$$
where DDS = 5-year diameter squared growth, BAL = basal area in larger trees,
SI = site index, and CR = crown ratio. FIAwrapper's `compute_competition()` and
`build_time_series()` produce all of these variables.
:::

### Mortality Modeling

```{r}
# Binary survival model
mort_data <- panel |>
  dplyr::filter(measurement_number > 1, !ingrowth) |>
  dplyr::mutate(died = as.integer(mortality))

mort_model <- glm(
  died ~ dbh + bal + plot_ba + spp_code,
  family = binomial,
  data = mort_data
)
summary(mort_model)
```

# Quality Control {#sec-quality}

## Outlier Detection {#sec-outliers}

`flag_outliers()` identifies trees with biologically implausible DBH, height,
or height-to-diameter ratios using species-group-specific bounds.

```{r}
trees <- build_tree_list(fia_ri, variant = "NE")
flagged <- flag_outliers(trees)

# How many warnings and flags?
sum(flagged$warn_dbh | flagged$warn_ht | flagged$warn_hd_ratio, na.rm = TRUE)
sum(flagged$flag_dbh | flagged$flag_ht | flagged$flag_hd_ratio, na.rm = TRUE)

# View flagged trees
flagged[!is.na(flagged$outlier_reason),
        c("spp_code", "dbh", "height", "outlier_reason")]
```

**Parameters**:

- `data`: Data frame with species, DBH, and height columns
- `spp_col`: Species code column name (default: `"spp_code"`)
- `dbh_col`: DBH column name (default: `"dbh"`)
- `ht_col`: Height column name (default: `"height"`)
- `method`: `"bounds"` (species-group biological limits) or `"iqr"` (1.5×IQR
  fences on H:D ratio within species groups)
- `sensitivity`: Multiplier for bounds (default 1.0; higher = less aggressive)

**Output columns added**:

| Column | Type | Description |
|--------|------|-------------|
| `warn_dbh` | logical | DBH exceeds 95th percentile for species group |
| `warn_ht` | logical | Height exceeds 95th percentile |
| `warn_hd_ratio` | logical | H:D ratio exceeds 95th percentile |
| `flag_dbh` | logical | DBH outside 99.5th percentile (likely error) |
| `flag_ht` | logical | Height outside 99.5th percentile |
| `flag_hd_ratio` | logical | H:D ratio outside 99.5th percentile |
| `outlier_reason` | character | Human-readable summary of all flags |

**Species groups** used for bounds:

| Group | Species Codes |
|-------|--------------|
| spruce_fir | BF, BS, RS, WS, NS, TA, TL |
| pine | WP, RP, JP, SP, LP |
| n_hardwood | SM, YB, AB, RO, WA, BA |
| cedar_hemlock | WC, NC, EH |
| o_hardwood | RM, QA, PB, GB, GA, BT, BP, SB, BC |
| other | All others |

Bounds are derived from FIA Phase 2 data (FIADB v9.2, northeastern evaluation
groups 2015–2023). All thresholds are in imperial units (DBH in inches, height
in feet).

::: {.callout-note}
## Flags vs Warnings
**Warnings** (95th percentile) indicate unusual but possible values — review
recommended. **Flags** (99.5th percentile) indicate values that are almost
certainly measurement errors or data entry mistakes. Neither removes data;
the user decides how to handle them.
:::

## Variance Estimation {#sec-variance}

`estimate_variance()` computes standard errors and confidence intervals for
plot-level attributes, supporting both simple random sampling (SRS) and
post-stratified estimation.

```{r}
plots <- get_plot_summary(trees)

# Simple random sampling estimate for BA/acre
est <- estimate_variance(
  stand = rep(1, nrow(plots)),
  plot = plots$plot_id,
  y = plots$ba_per_acre,
  conf_level = 0.90
)

est$mean           # Mean BA/acre
est$se             # Standard error
est$ci_lower       # 90% CI lower bound
est$ci_upper       # 90% CI upper bound
est$precision_pct  # Precision as % of mean (= SE/mean × 100)
est$n_plots        # Number of plots
est$method         # "SRS" or "post_stratified"
```

**Post-stratified estimation** (reduces variance when strata are available):

```{r}
# Stratify by forest type
site <- get_site_data(fia_ri)
plot_data <- merge(plots, site[, c("plot_cn", "forest_type_code")], by = "plot_cn")

est_strat <- estimate_variance(
  stand = rep(1, nrow(plot_data)),
  plot = plot_data$plot_id,
  y = plot_data$ba_per_acre,
  stratum = plot_data$forest_type_code,
  stratum_weight = NULL,  # uses proportional weights when NULL
  conf_level = 0.90
)
```

**References**:

- Bechtold & Patterson (2005). *The Enhanced FIA Program*. Gen. Tech. Rep. SRS-80.
- Scott et al. (2005). *Sample-Based Estimators Used by the Forest Inventory and
  Analysis National Information Management System*. GTR-SRS-80, Ch. 4.

## Regeneration Assessment {#sec-regen}

`summarize_regen()` computes regeneration metrics from microplot data (saplings
with DBH < 5.0 inches).

```{r}
regen <- summarize_regen(
  stand   = trees$plot_id,
  plot    = trees$plot_cn,
  subplot = trees$subplot,
  dbh     = trees$dbh,
  species = trees$spp_code
)

head(regen[, c("stand", "plot", "regen_tpa", "regen_ba_per_acre",
               "stocking_pct", "n_regen_trees")])
```

**Output columns**:

| Column | Units | Description |
|--------|-------|-------------|
| `regen_tpa` | trees/acre | Regeneration density |
| `regen_ba_per_acre` | ft²/acre | Regeneration basal area |
| `stocking_pct` | percent | Stocking relative to reference density |
| `n_regen_trees` | count | Number of regeneration trees measured |
| `n_subplots` | count | Number of subplots with data |
| `commercial_regen_tpa` | trees/acre | TPA of commercial species (if specified) |

The function filters to trees with DBH < 5.0 inches (the microplot threshold)
and computes per-acre expansion using the 6.8-foot microplot radius.

**Commercial species filtering**:

```{r}
# Only count commercially valuable species
regen_comm <- summarize_regen(
  stand = trees$plot_id,
  plot = trees$plot_cn,
  subplot = trees$subplot,
  dbh = trees$dbh,
  species = trees$spp_code,
  commercial_species = c("WP", "RO", "SM", "WO")
)
```

# Species Code System {#sec-species}

FIA uses numeric species codes (SPCD), while FVS uses 2-letter alpha codes that
vary by regional variant. FIAwrapper handles all translations automatically, but
you can also use the species functions directly.

**References**:

- [FIA REF_SPECIES table](https://apps.fs.usda.gov/fia/datamart/CSV/REF_SPECIES.csv)
  — Complete FIA species list
- [FVS Species Codes by Variant](https://www.fs.usda.gov/fvs/software/plantnat.php)
  — Official FVS species mappings

## Species Code Functions

### `fia_to_fvs()` — FIA Numeric → FVS Alpha

```{r}
fia_to_fvs(c(316, 833, 129, 261, 12))
#> [1] "RM" "RO" "WP" "EH" "BF"

# Unknown species get "OH" (other hardwood)
fia_to_fvs(99999)
#> [1] "OH"
```

### `fvs_to_fia()` — FVS Alpha → FIA Numeric

```{r}
fvs_to_fia(c("RM", "RO", "WP", "EH", "BF"))
#> [1] 316 833 129 261  12

# Unknown codes get 999
fvs_to_fia("ZZ")
#> [1] 999
```

### `get_common_name()` — FIA Code → Common Name

```{r}
get_common_name(c(316, 833, 129, 261, 12))
#> [1] "red maple"          "red oak"            "eastern white pine"
#> [4] "eastern hemlock"    "balsam fir"
```

### `get_species_table()` — Full Mapping Table

```{r}
spp_table <- get_species_table("NE")
head(spp_table)
#>   spcd fvs_code       common_name        scientific_name wood_type
#> 1   12       BF       balsam fir        Abies balsamea  softwood
#> 2   68       RC  eastern redcedar  Juniperus virginiana  softwood
#> ...
```

The table includes `spcd`, `fvs_code`, `common_name`, `scientific_name`, and
`wood_type` ("hardwood" or "softwood") for all species in the variant.

### `pef_to_fvs()` — Penobscot Experimental Forest Codes

```{r}
# For users working with PEF data
pef_to_fvs(c(1, 2, 3, 4))
#> [1] "BF" "RS" "WS" "RS"
```

## Variant-Aware Lookups

Species codes are **variant-aware**: the same SPCD may map to different FVS
codes depending on the regional variant. FIAwrapper loads the correct mapping
table from `inst/extdata/fvs_species_variants/` based on the active variant.

```{r}
# NE variant
set_fvs_variant("NE")
fia_to_fvs(316)  # red maple
#> [1] "RM"

# SN variant
set_fvs_variant("SN")
fia_to_fvs(316)  # same species, might have different code
#> [1] "RM"
```

Currently bundled variant tables: **NE** (75 species), **SN** (61 species),
**LS** (42 species). Additional variants fall back to the default table.

# Utility Functions {#sec-utilities}

## Basal Area Computation

```{r}
# Basal area in square feet (from DBH in inches)
ba_ft2(10)
#> [1] 0.5454154

# Formula: DBH^2 × 0.005454154 = DBH^2 × (π/4) / 144
ba_ft2(c(8, 12, 16, 20))
#> [1] 0.3491 0.7854 1.3963 2.1817

# Metric: basal area in square meters (from DBH in cm)
ba_m2(25.4)
#> [1] 0.05067
```

## Quadratic Mean Diameter

```{r}
qmd(
  dbh = c(8, 10, 12, 14, 16),
  tpa = c(20, 15, 12, 8, 5)
)
#> [1] 11.15
```

$QMD = \sqrt{\frac{\sum DBH_i^2 \times TPA_i}{\sum TPA_i}}$

## TPA Expansion Factors

```{r}
# Compute TPA from first principles
compute_tpa(
  dbh = c(12, 8, 3.5, 14),
  subplot = c(1, 2, 3, 4),
  cond_prop = c(1.0, 1.0, 1.0, 0.8)  # last tree in split condition
)
#> [1]  6.018  6.018 74.965  7.523
```

**Parameters**:

- `dbh`: DBH in inches (determines subplot vs microplot)
- `subplot`: Subplot number (used to count unique subplots)
- `cond_prop`: Condition proportion (optional, adjusts TPA)
- `subplot_radius_ft`: Default 24.0 ft
- `microplot_radius_ft`: Default 6.8 ft
- `dbh_threshold`: Default 5.0 inches

## Column Name Translation

FIAwrapper provides bidirectional column name translation between FIA database
names and forestry terms.

### `fia_to_forestry_names()`

```{r}
df <- data.frame(DIA = 10.5, HT = 65, STATUSCD = 1, SPCD = 316)
fia_to_forestry_names(df)
#>    dbh height status_code spcd
#> 1 10.5     65           1  316
```

### `forestry_to_fia_names()`

```{r}
df <- data.frame(dbh = 10.5, height = 65, status_code = 1)
forestry_to_fia_names(df)
#>    DIA  HT STATUSCD
#> 1 10.5  65        1
```

**Full translation mapping** — see @sec-column-ref.

## Unit Conversions (Internal Helpers)

These are available but primarily used internally:

```{r}
inches_to_cm(10)    # 25.4
cm_to_inches(25.4)  # 10
feet_to_meters(65)  # 19.812
meters_to_feet(20)  # 65.617
acres_to_hectares(1)    # 0.4047
hectares_to_acres(0.5)  # 1.2355
```

# Spatial API Functions {#sec-spatial-apis}

FIAwrapper can enrich plot data with terrain, climate, and soils information
from federal web APIs. These functions require an `sf` POINT object (from
`plot_locations()`) and network access.

## Terrain Data (USGS 3DEP) {#sec-terrain}

`get_terrain_data()` fetches high-resolution elevation from the
[USGS 3D Elevation Program](https://www.usgs.gov/3d-elevation-program) (3DEP)
National Map API.

```{r}
plots_sf <- plot_locations(fia_ri)
plots_terrain <- get_terrain_data(plots_sf)

head(plots_terrain[, c("plot_cn", "dem_elevation")])
```

**Returns**: The input `sf` object with an added `dem_elevation` column (feet).
This is DEM-derived elevation at ~10m resolution, more precise than FIA's
recorded `ELEV` value.

**API reference**: [USGS EPQS](https://epqs.nationalmap.gov/v1/docs)

## Climate Data (Daymet) {#sec-climate}

`get_climate_data()` fetches annual climate summaries from
[Daymet](https://daymet.ornl.gov/), a 1-km gridded climate dataset produced by
Oak Ridge National Laboratory.

```{r}
plots_climate <- get_climate_data(plots_sf, years = 2019)

head(plots_climate[, c("plot_cn", "mean_annual_temp", "annual_precip")])
```

**Returns**: Added columns `mean_annual_temp` (°F) and `annual_precip` (inches).

**Parameters**:

- `years`: Integer year(s) to fetch (default: 2020)
- `source`: Currently `"daymet"` only

**API reference**: [Daymet Single Pixel](https://daymet.ornl.gov/web_services)

**Climate variables available from Daymet** (FIAwrapper retrieves temp and
precip; additional variables available via direct API access):

| Variable | Description |
|----------|-------------|
| tmax, tmin | Daily max/min temperature (°C) |
| prcp | Daily precipitation (mm) |
| srad | Shortwave radiation (W/m²) |
| vp | Vapor pressure (Pa) |
| swe | Snow water equivalent (kg/m²) |
| dayl | Day length (seconds) |

## Soils Data (SSURGO) {#sec-soils}

`get_soils_data()` queries the
[USDA SSURGO database](https://www.nrcs.usda.gov/resources/data-and-reports/soil-survey-geographic-database-ssurgo)
via the Soil Data Access (SDA) web API.

```{r}
plots_soils <- get_soils_data(plots_sf)

head(plots_soils[, c("plot_cn", "soil_texture", "drainage_class", "awc", "soil_depth")])
```

**Returns**: Added columns:

| Column | Description |
|--------|-------------|
| `soil_texture` | Soil texture class (e.g., "sandy loam", "silt loam") |
| `drainage_class` | Drainage class (e.g., "well drained", "poorly drained") |
| `awc` | Available water capacity (inches/inch) |
| `soil_depth` | Depth to restrictive layer (inches) |

**API reference**: [SDA Query](https://sdmdataaccess.nrcs.usda.gov/)

::: {.callout-tip}
## Combining Spatial Data for Modeling
```{r}
# Build a complete modeling dataset
plots_sf <- plot_locations(fia_ri)
plots_sf <- get_terrain_data(plots_sf)
plots_sf <- get_climate_data(plots_sf, years = 2019)
plots_sf <- get_soils_data(plots_sf)

# Join to tree data for individual tree models
trees <- build_tree_list(fia_ri, variant = "NE")
# merge on plot_cn...
```
:::

# End-to-End Workflows {#sec-workflows}

## Workflow 1: Property Inventory → FVS {#sec-workflow-property}

A consulting forester receives a shapefile of a client's property and needs to
produce FVS-ready stand inventories.

```{r}
library(FIAwrapper)
set_fvs_variant("NE")
data(fia_ri)  # substitute with real state data

# ---- Step 1: Build tree list ----
# In practice: get_tree_list(states = "ME", aoi = "client.shp")
trees <- build_tree_list(fia_ri, include_dead = TRUE, variant = "NE")
cat("Trees:", nrow(trees), "on", length(unique(trees$plot_id)), "plots\n")

# ---- Step 2: Quality check ----
flagged <- flag_outliers(trees)
issues <- flagged[!is.na(flagged$outlier_reason), ]
cat(nrow(issues), "trees flagged for review\n")

# Review flagged trees
if (nrow(issues) > 0) {
  print(issues[, c("plot_id", "spp_code", "dbh", "height", "outlier_reason")])
}

# ---- Step 3: Plot summaries ----
plots <- get_plot_summary(trees[trees$status == "live", ])
cat("\nPlot Summary Statistics:\n")
cat("  Mean TPA:", round(mean(plots$tpa), 1), "trees/acre\n")
cat("  Mean BA/ac:", round(mean(plots$ba_per_acre), 1), "ft²/acre\n")
cat("  Mean QMD:", round(mean(plots$qmd, na.rm = TRUE), 1), "inches\n")

# ---- Step 4: Species composition ----
spp <- get_species_summary(trees[trees$status == "live", ])
top_spp <- spp |>
  dplyr::group_by(spp_code, species_name) |>
  dplyr::summarize(mean_ba = round(mean(ba_per_acre), 1), .groups = "drop") |>
  dplyr::arrange(dplyr::desc(mean_ba))
cat("\nTop Species by BA/acre:\n")
print(head(top_spp, 8))

# ---- Step 5: Site data ----
site <- get_site_data(fia_ri)

# ---- Step 6: Export to FVS ----
export_fvs_treelist(
  trees[trees$status == "live", ],
  "client_fvs.tre",
  variant = "NE",
  format = "standard"
)

stand_data <- merge(plots, site, by = "plot_cn")
export_fvs_standlist(stand_data, "client_stands.csv", variant = "NE")

cat("\nFVS files exported. Ready for growth projection.\n")
```

## Workflow 2: Regional Growth Analysis {#sec-workflow-growth}

A researcher wants to build diameter increment models from FIA remeasurement
data across a state.

```{r}
library(FIAwrapper)
set_fvs_variant("NE")
data(fia_ri)  # substitute: fetch_fia_data("ME", data_dir = "~/fia")

# ---- Step 1: Identify remeasured plots ----
remeas <- get_remeasured_plots(fia_ri)
cat(nrow(remeas), "plots with multiple measurements\n")
cat("Year range:", min(remeas$first_year), "-", max(remeas$last_year), "\n")

# ---- Step 2: Build time series ----
ts <- build_time_series(fia_ri, include_dead = TRUE, variant = "NE")
cat(nrow(ts), "tree-measurement records\n")

# ---- Step 3: Plot-level growth summary ----
growth <- summarize_growth(ts)
cat("\nPlot-Level Growth Summary:\n")
cat("  Mean net BA growth:", round(mean(growth$net_ba_growth_per_acre, na.rm = TRUE), 2),
    "ft²/ac/yr\n")
cat("  Mean mortality rate:", round(mean(growth$mortality_rate, na.rm = TRUE) * 100, 1),
    "%\n")

# ---- Step 4: Individual tree panel data ----
panel <- get_individual_tree_history(ts, min_measurements = 2)

# Filter to quality growth records
model_data <- panel |>
  dplyr::filter(
    status == "live",
    !ingrowth,
    !is.na(annual_dbh_growth),
    annual_dbh_growth > 0,
    annual_dbh_growth < 1.0
  )
cat(nrow(model_data), "growth records for modeling\n")

# ---- Step 5: Fit diameter increment model ----
mod <- lm(
  log(annual_dbh_growth) ~ log(dbh) + bal + spp_code,
  data = model_data
)

cat("\nDiameter Increment Model:\n")
cat("  R² =", round(summary(mod)$r.squared, 3), "\n")
cat("  n  =", nrow(model_data), "\n")

# ---- Step 6: Summarize growth by species ----
growth_by_spp <- model_data |>
  dplyr::group_by(spp_code) |>
  dplyr::summarize(
    n_trees = dplyr::n(),
    mean_dbh = round(mean(dbh), 1),
    mean_growth = round(mean(annual_dbh_growth), 3),
    sd_growth = round(sd(annual_dbh_growth), 3),
    .groups = "drop"
  ) |>
  dplyr::arrange(dplyr::desc(n_trees))

cat("\nGrowth by Species:\n")
print(growth_by_spp)
```

## Workflow 3: Species Composition Change Over Time {#sec-workflow-species}

Tracking how species composition shifts between inventory cycles.

```{r}
library(FIAwrapper)
set_fvs_variant("NE")
data(fia_ri)

# ---- Step 1: Build time series (live trees only) ----
ts <- build_time_series(fia_ri, include_dead = FALSE, variant = "NE")

# ---- Step 2: Species BA by measurement period ----
spp_time <- ts |>
  dplyr::group_by(inventory_year, spp_code) |>
  dplyr::summarize(
    total_ba = sum(ba_per_acre, na.rm = TRUE),
    total_tpa = sum(tpa, na.rm = TRUE),
    n_plots = dplyr::n_distinct(plot_id),
    .groups = "drop"
  ) |>
  dplyr::mutate(
    mean_ba_per_acre = total_ba / n_plots,
    mean_tpa = total_tpa / n_plots
  )

# ---- Step 3: Compare hardwood vs softwood ----
spp_table <- get_species_table("NE")
spp_time <- dplyr::left_join(
  spp_time,
  spp_table[, c("fvs_code", "wood_type")],
  by = c("spp_code" = "fvs_code")
)

wood_trend <- spp_time |>
  dplyr::group_by(inventory_year, wood_type) |>
  dplyr::summarize(
    ba_per_acre = sum(mean_ba_per_acre, na.rm = TRUE),
    .groups = "drop"
  )

cat("Hardwood vs Softwood BA/acre by Year:\n")
print(wood_trend)

# ---- Step 4: Top species trends ----
top_spp_codes <- spp_time |>
  dplyr::group_by(spp_code) |>
  dplyr::summarize(total = sum(mean_ba_per_acre)) |>
  dplyr::arrange(dplyr::desc(total)) |>
  head(5) |>
  dplyr::pull(spp_code)

spp_trend <- spp_time |>
  dplyr::filter(spp_code %in% top_spp_codes) |>
  dplyr::select(inventory_year, spp_code, mean_ba_per_acre)

cat("\nTop 5 Species BA/acre Trends:\n")
print(tidyr::pivot_wider(spp_trend, names_from = inventory_year,
                          values_from = mean_ba_per_acre))
```

## Workflow 4: Site Quality Assessment {#sec-workflow-site}

Evaluating site quality across plots using FIA attributes and spatial API data.

```{r}
library(FIAwrapper)
set_fvs_variant("NE")
data(fia_ri)

# ---- Step 1: Build tree list and summaries ----
trees <- build_tree_list(fia_ri, variant = "NE")
plots <- get_plot_summary(trees)
site  <- get_site_data(fia_ri)

# ---- Step 2: Combine into site assessment table ----
assessment <- merge(plots, site, by = "plot_cn", suffixes = c("", "_site"))

# Key site quality indicators
cat("Site Quality Assessment:\n")
cat("  Site Index range:", min(assessment$site_index, na.rm = TRUE),
    "-", max(assessment$site_index, na.rm = TRUE), "ft\n")
cat("  Stand Age range:", min(assessment$stand_age, na.rm = TRUE),
    "-", max(assessment$stand_age, na.rm = TRUE), "years\n")
cat("  BA/acre range:", round(min(assessment$ba_per_acre), 1),
    "-", round(max(assessment$ba_per_acre), 1), "ft²/ac\n")

# ---- Step 3: Enrich with spatial data (requires network) ----
# plots_sf <- plot_locations(fia_ri)
# plots_sf <- get_terrain_data(plots_sf)
# plots_sf <- get_climate_data(plots_sf, years = 2019)
# plots_sf <- get_soils_data(plots_sf)

# ---- Step 4: Site index vs BA relationship ----
cat("\nSite Index vs BA/acre correlation:",
    round(cor(assessment$site_index, assessment$ba_per_acre,
              use = "complete.obs"), 3), "\n")

# ---- Step 5: Variance estimate for the area ----
est <- estimate_variance(
  stand = rep(1, nrow(assessment)),
  plot = assessment$plot_id,
  y = assessment$ba_per_acre,
  conf_level = 0.90
)
cat("\nBA/acre estimate:\n")
cat("  Mean:", round(est$mean, 1), "ft²/ac\n")
cat("  SE:", round(est$se, 1), "ft²/ac\n")
cat("  90% CI:", round(est$ci_lower, 1), "-", round(est$ci_upper, 1), "ft²/ac\n")
cat("  Precision:", round(est$precision_pct, 1), "% of mean\n")
```

# Appendices {.unnumbered}

## Appendix A: Complete Column Name Reference {#sec-column-ref}

The `fia_to_forestry_names()` and `forestry_to_fia_names()` functions use
this mapping:

### Tree Table Columns

| FIA Column | FIAwrapper Name | Description |
|-----------|----------------|-------------|
| CN | tree_cn | Unique tree control number |
| PLT_CN | plot_cn | Plot control number |
| CONDID | condition_id | Condition ID within plot |
| SUBP | subplot | Subplot number (1–4) |
| TREE | tree_num | Tree number within subplot |
| STATUSCD | status_code | 1=live, 2=dead, 3=removed |
| SPCD | spcd | FIA numeric species code |
| DIA | dbh | Diameter at breast height (inches) |
| HT | height | Total tree height (feet) |
| ACTUALHT | actual_height | Actual measured height (feet) |
| CR | crown_ratio | Compacted crown ratio (%) |
| CCLCD | crown_class | Crown class code |
| TPA_UNADJ | tpa_unadj | FIA unadjusted TPA |
| VOLCFNET | vol_cf_net | Net cubic foot volume |
| VOLCFGRS | vol_cf_gross | Gross cubic foot volume |
| VOLBFNET | vol_bf_net | Net board foot volume (Scribner) |
| DRYBIO_AG | biomass_ag_dry | Aboveground dry biomass (lbs) |
| CARBON_AG | carbon_ag | Aboveground carbon (lbs) |
| PREV_TRE_CN | prev_tree_cn | CN at previous measurement |
| DECAYCD | decay_class | Decay class (dead trees, 1–5) |
| DSTRBCD1 | disturbance_code1 | Primary disturbance code |
| DAMAGE_AGENT_CD1 | damage_code1 | Primary damage agent code |

### Plot Table Columns

| FIA Column | FIAwrapper Name | Description |
|-----------|----------------|-------------|
| CN | plot_cn | Unique plot control number |
| STATECD | state_code | FIPS state code |
| COUNTYCD | county_code | FIPS county code |
| PLOT | plot_number | Plot number within county |
| INVYR | inventory_year | Inventory year |
| LAT | lat | Latitude (fuzzed, WGS84) |
| LON | lon | Longitude (fuzzed, WGS84) |
| ELEV | elevation | Elevation (feet) |
| ECOSUBCD | ecological_subsection | Ecological subsection code |

### Condition Table Columns

| FIA Column | FIAwrapper Name | Description |
|-----------|----------------|-------------|
| CONDPROP_UNADJ | cond_proportion | Proportion of plot in condition (0–1) |
| FORTYPCD | forest_type_code | FIA forest type code |
| STDAGE | stand_age | Stand age (years) |
| SICOND | site_index | Site index (base age 50, feet) |
| SITECLCD | site_class | Site productivity class (1–7) |
| SLOPE | slope | Slope (percent) |
| ASPECT | aspect | Aspect (degrees) |
| BALIVE | ba_live | Live BA on condition (ft²/ac) |
| SDI | stand_density_index | Stand density index |
| OWNGRPCD | ownership_group | Ownership group code |
| PHYSCLCD | physiographic_class | Physiographic class code |
| DSTRBCD1 | disturbance_code1 | Primary disturbance code |

## Appendix B: FIA Forest Type Codes {#sec-forest-types}

Common forest type codes found in the COND table `FORTYPCD` field. Full list:
[REF_FOREST_TYPE](https://apps.fs.usda.gov/fia/datamart/CSV/REF_FOREST_TYPE.csv).

| Code | Forest Type | Region |
|------|------------|--------|
| 101 | Jack pine | Lake States, NE |
| 102 | Red spruce/balsam fir | NE |
| 103 | Eastern white pine/eastern hemlock | NE |
| 104 | White pine | NE, Mid-Atlantic |
| 105 | Red pine | NE, Lake States |
| 108 | Norway spruce | NE |
| 301 | Scotch pine | Various |
| 401 | Sugar maple/beech/yellow birch | NE |
| 402 | Black cherry | NE, Mid-Atlantic |
| 403 | White oak/red oak/hickory | Eastern US |
| 501 | White oak | Eastern US |
| 503 | Northern red oak | NE, Mid-Atlantic |
| 505 | Scarlet oak | SE, Mid-Atlantic |
| 509 | Red maple/oak | NE |
| 511 | Red maple/lowland | NE |
| 520 | Mixed hardwoods | Various |
| 701 | Eastern white pine | NE |
| 801 | White spruce | NE, Lake States |
| 802 | Black spruce | NE, Lake States |
| 805 | Balsam fir | NE |
| 901 | Atlantic white-cedar | NE coast |

## Appendix C: FIA Status Codes and Damage Agents {#sec-status-codes}

### Tree Status (STATUSCD)

| Code | Status | Description |
|------|--------|-------------|
| 1 | Live | Tree is alive |
| 2 | Dead | Standing dead tree (snag) |
| 3 | Removed | Tree was removed (harvested, cleared) |

### Site Productivity Class (SITECLCD)

| Code | Board Feet/Acre/Year | Description |
|------|---------------------|-------------|
| 1 | 225+ | Very high productivity |
| 2 | 165–224 | High productivity |
| 3 | 120–164 | Moderate-high productivity |
| 4 | 85–119 | Moderate productivity |
| 5 | 50–84 | Low-moderate productivity |
| 6 | 20–49 | Low productivity |
| 7 | 0–19 | Very low productivity |

### Crown Class (CCLCD)

| Code | Crown Class | Description |
|------|------------|-------------|
| 1 | Open-grown | Free from competition |
| 2 | Dominant | Crown extends above general canopy |
| 3 | Codominant | Crown in general canopy level |
| 4 | Intermediate | Crown below but extending into canopy |
| 5 | Overtopped | Crown entirely below canopy |

## Appendix D: Function Quick Reference {#sec-function-ref}

### Data Acquisition

| Function | Purpose |
|----------|---------|
| `fetch_fia_data(states, tables, data_dir, method)` | Download FIA data |
| `load_fia_data(data_dir, states)` | Load cached FIA data |

### Spatial

| Function | Purpose |
|----------|---------|
| `get_plots_in_area(aoi, fia_data, buffer_ft)` | Filter plots to AOI |
| `plot_locations(fia_data, crs)` | Create sf POINT from plots |
| `map_plots(fia_data, aoi)` | Quick interactive map |
| `get_terrain_data(plots_sf)` | USGS 3DEP elevation |
| `get_climate_data(plots_sf, years)` | Daymet climate data |
| `get_soils_data(plots_sf)` | SSURGO soil data |

### Tree Lists

| Function | Purpose |
|----------|---------|
| `build_tree_list(fia_data, include_dead, variant)` | Build clean tree list |
| `get_tree_list(states, aoi, include_dead, variant)` | One-line download + build |
| `prep_fia_data(fia_trees, fia_plots, fia_cond)` | Low-level data prep |

### Summaries

| Function | Purpose |
|----------|---------|
| `get_plot_summary(tree_list, include_dead)` | Per-plot aggregation |
| `get_site_data(fia_data)` | Site attributes from PLOT + COND |
| `get_species_summary(tree_list)` | Per-species-per-plot summary |
| `summarize_plot(stand, plot, dbh, height, tpa)` | Vector-based plot summary |

### Time Series

| Function | Purpose |
|----------|---------|
| `get_remeasured_plots(fia_data)` | Find multi-cycle plots |
| `build_time_series(fia_data, include_dead, variant)` | Link trees across cycles |
| `summarize_growth(time_series)` | Plot-level growth metrics |
| `get_individual_tree_history(time_series)` | Panel data for modeling |
| `compute_competition(tree_list)` | BAL, relative DBH, plot BA |

### FVS Export

| Function | Purpose |
|----------|---------|
| `set_fvs_variant(variant)` | Set session FVS variant |
| `export_fvs_treelist(tree_list, file, variant, format)` | Write FVS tree file |
| `export_fvs_standlist(plot_summary, file, variant)` | Write FVS stand file |

### Species Codes

| Function | Purpose |
|----------|---------|
| `fia_to_fvs(spcd, variant)` | FIA numeric → FVS alpha |
| `fvs_to_fia(fvs_code, variant)` | FVS alpha → FIA numeric |
| `get_common_name(spcd)` | FIA code → common name |
| `get_species_table(variant)` | Full mapping table |
| `pef_to_fvs(pef_code)` | PEF code → FVS alpha |

### Quality Control

| Function | Purpose |
|----------|---------|
| `flag_outliers(data, method, sensitivity)` | Detect outlier trees |
| `estimate_variance(stand, plot, y, stratum)` | SE and confidence intervals |
| `summarize_regen(stand, plot, subplot, dbh)` | Regeneration metrics |

### Utilities

| Function | Purpose |
|----------|---------|
| `ba_ft2(dbh_inches)` | Basal area (ft²) |
| `ba_m2(dbh_cm)` | Basal area (m²) |
| `qmd(dbh, tpa)` | Quadratic mean diameter |
| `compute_tpa(dbh, subplot, cond_prop)` | Per-tree expansion factor |
| `fia_to_forestry_names(df)` | FIA → forestry column names |
| `forestry_to_fia_names(df)` | Forestry → FIA column names |
| `check_python_deps()` | Verify Python environment |
| `ensure_python_env()` | Create/update Python environment |

## Appendix E: External Resources {#sec-resources}

### FIA Program

- [FIA National Program](https://www.fia.fs.usda.gov/) — Official website
- [FIADB User Guide v9.2](https://www.fia.fs.usda.gov/library/database-documentation/)
  — Complete database documentation (essential reference)
- [FIA DataMart](https://apps.fs.usda.gov/fia/datamart/) — Download raw FIA data
- [EVALIDator](https://apps.fs.usda.gov/fiadb-api/evalidator) — Web-based
  estimation tool (use to verify FIAwrapper outputs)
- [FIA Spatial Data](https://www.fia.fs.usda.gov/tools-data/spatial/index.php)
  — Spatial data request portal

### FVS (Forest Vegetation Simulator)

- [FVS Overview](https://www.fs.usda.gov/fvs/whatis.shtml) — What is FVS
- [FVS Variants](https://www.fs.usda.gov/fvs/software/plantnat.php) — Regional
  models and species coverage
- [Suppose User Guide](https://www.fs.usda.gov/fvs/documents/userguides.shtml)
  — GUI for running FVS
- [FVS Species Codes](https://www.fs.usda.gov/fvs/documents/gtrs.shtml) — Alpha
  codes by variant

### R Packages for Forestry

- [rFIA](https://cran.r-project.org/package=rFIA) — R interface to FIA data
  (Hunter Stanke). Used by FIAwrapper for data downloads.
- [ForestTools](https://cran.r-project.org/package=ForestTools) — Remote sensing
  for forestry
- [lmfor](https://cran.r-project.org/package=lmfor) — Forest biometrics models

### Federal APIs Used by FIAwrapper

- [USGS 3DEP / EPQS](https://www.usgs.gov/3d-elevation-program) — Elevation
  data (10m resolution)
- [Daymet](https://daymet.ornl.gov/) — Gridded daily climate (1km resolution,
  ORNL DAAC)
- [SSURGO / SDA](https://www.nrcs.usda.gov/resources/data-and-reports/soil-survey-geographic-database-ssurgo)
  — Soil survey data (USDA NRCS)

### Key Literature

- Bechtold & Patterson (2005). *The Enhanced FIA Program — National Consistency
  with Regional Relevance*. Gen. Tech. Rep. SRS-80.
- Burrill et al. (2024). *The Forest Inventory and Analysis Database: Database
  Description and User Guide for Phase 2*. Version 9.2.
- Dixon (2002). *Essential FVS: A User's Guide to the Forest Vegetation
  Simulator*. USDA-FS Internal Report.
- Wykoff (1990). *A Basal Area Increment Model for Individual Conifers in the
  Northern Rocky Mountains*. Forest Science 36(4):1077-1104.
- Monserud & Sterba (1999). *Modeling Individual Tree Mortality for Austrian
  Forest Species*. Forest Ecology and Management 113:109-123.
- Scott et al. (2005). *Sample-Based Estimators Used by the Forest Inventory
  and Analysis National Information Management System*. GTR-SRS-80, Chapter 4.
- Shaw (2008). *Evaluation and Use of FIA Data for Growth and Yield
  Applications*. GTR-NRS-P-56.
- Reineke (1933). *Perfecting a Stand-Density Index for Even-Aged Forests*.
  Journal of Agricultural Research 46:627-638.
